A novel mechanism for bacterial sporulation based on programmed peptidoglycan degradation

Version 1

Abstract

Abstract

Many bacteria form spores to endure unfavorable conditions. While Firmicutes generate endospores through cell division, sporulation in non-Firmicutes remains less understood. The Gram-negative bacterium Myxococcus xanthus undergoes sporulation through two distinct mechanisms: rapid sporulation triggered by chemical induction and slow sporulation driven by starvation, both occurring independently of cell division. Instead, these processes depend on the complete degradation of the peptidoglycan (PG) cell wall by lytic transglycosylases (LTGs), with both LtgA and LtgB supporting rapid sporulation and LtgB alone driving slow sporulation. Remarkably, LtgB programs the pace of PG degradation by LtgA during rapid sporulation, ensuring a controlled process that prevents abrupt PG breakdown and the formation of non-resistant pseudospores. In addition to regulation between LTGs, PG degradation is also influenced by its synthesis; cells exhibiting increased muropeptide production often circumvent sporulation. These findings not only reveal novel mechanisms of bacterial sporulation but also shed light on the regulatory network governing PG dynamics.

Introduction

Spores are metabolically dormant cells that can survive unfavorable conditions, including extremes of temperature, desiccation, and ionizing radiation (Huang & Hull, 2017, Hutchison et al., 2014). Sporulation, the process of spore development, is a strategy utilized by a wide variety of organisms, from bacteria and protozoa to fungi and plants. In addition to protective spore coats, spore resilience relies on cytosol dehydration and DNA compaction—key processes that decrease cell volume and alter cell shape (Higgins & Dworkin, 2012). Thus, morphological differentiation is a hallmark of sporulation. In bacterial spore formers, as peptidoglycan (PG) cell walls largely determine cell morphology, their sporulation requires profound PG remodeling.

PG is a rigid, mesh-like macromolecule that is composed of glycan strands of repeating units of N-acetyl glucosamine (GluNAc)-N-acetyl muramic acid (MurNAc) crosslinked by peptides (Egan et al., 2020). PG encloses the entire cytoplasmic membrane, and its rigidity provides mechanical support against osmotic stress. For this reason, PG is an essential structure for most bacteria and a major target for antibacterial treatments. Sporulation provides invaluable opportunities for understanding the dynamics of PG, which is controlled by multiple synthases and hydrolases. PG synthases include glycosyltransferases (GTases) that polymerize glycan strands and transpeptidases (TPases) that form peptide crosslinks (Egan et al., 2020). PG hydrolases, also known as autolysins, include glycosidases that break the glycan strands, and amidases and endo/carboxypeptidases that cleave the peptide crosslinks (Egan et al., 2020, Rohs & Bernhardt, 2021, van Heijenoort, 2011).

Firmicutes, including Gram-positive bacteria like Bacilli and Clostridia, produce endospores. Their morphological transition from rod-shaped vegetative cells to ovoid spores occurs through asymmetric cell division, resulting in the formation of a smaller forespore and a larger mother cell. Eventually, the forespore becomes an oval endospore after being engulfed by the mother cell (Higgins & Dworkin, 2012, McKenney & Eichenberger, 2012). A mature endospore contains two PG layers: the germ cell wall, derived from the forespore’s original PG, and a thickened PG cortex deposited by the mother cell (Popham & Bernhards, 2015). While a few Gram-negative spore formers also belong to the Firmicutes phylum, they share conserved sporulation genes with Bacilli and Clostridia, which suggests similar sporulation mechanisms (Tocheva et al., 2011, Yutin & Galperin, 2013).

Sporulation by non-firmicutes bacteria has been largely overlooked. Myxobacteria, a group of Gram-negative, non-firmicutes spore formers, were first assigned to the phylum δ-proteobacteria, but recently reclassified into the newly established phylum Myxococcota (Waite et al., 2020). Myxococcus xanthus is a model organism of myxobacteria. Rod-shaped M. xanthus cells can undergo sporulation via two distinct pathways, both resulting in spherical spores. First, in response to certain chemical signals, such as glycerol and dimethyl sulfoxide (DMSO), individual M. xanthus cells can rapidly transition into isolated spherical spores in aqueous environments within hours (Dworkin & Gibson, 1964). Second, millions of cells can aggregate on solid starvation media and develop into spore-filled fruiting bodies, a process that takes a few days to complete (O’Connor & Zusman, 1988). Both sporulation pathways are programmed, tightly controlled processes that involve over 1,000 genes (Muller et al., 2010, Munoz-Dorado et al., 2019). In contrast to endospore formation, cell division is not involved in either the M. xanthus sporulation mechanisms, and M. xanthus lacks the sporulation genes in Firmicutes (Aramayo & Nan, 2022). Thus, sporulating M. xanthus must accomplish the rod-to-sphere transition through yet to be discovered PG remodeling mechanisms, which provide invaluable opportunities to investigate PG dynamics (Zhang et al., 2021).

Distinct from Bacilli and Clostridia, M. xanthus spores lack cortex PG (Bui et al., 2009, Zhang et al., 2020, Voelz & Dworkin, 1962). Rather, their resistance is derived from polysaccharide coats (Kottel et al., 1975, Perez-Burgos et al., 2020, Saidi et al., 2021, Muller et al., 2012). Sporulating cells that fail to deposit coat polysaccharides on their surfaces produce spherical pseudospores that lack resistance (Wartel et al., 2013, Zhang et al., 2020, Holkenbrink et al., 2014). In a pioneering study, Bui et al. did not detect muropeptides in glycerol-induced spores, indicating that M. xanthus degrades its PG during rapid sporulation (Bui et al., 2009). However, it remains unclear whether starvation-induced spores within fruiting bodies still retain PG and how M. xanthus orchestrates programmed morphological changes.

In this report, we used ultra-performance liquid chromatography (UPLC) to demonstrate that M. xanthus degrades PG in both sporulation pathways. Through mutagenesis studies, we discovered that these sporulation pathways require different lytic transglycosylases (LTGs) for PG degradation. While the rapid sporulation relies on both LtgA and LtgB, the slow formation of fruiting body spores only requires LtgB. Remarkably, LtgB regulates the pace of PG degradation by LtgA during rapid sporulation, preventing the formation of non-resistant pseudospores due to abrupt PG breakdown. This research not only uncovers novel mechanisms of sporulation in non-firmicutes but also highlights the crucial role of cross-regulation between PG hydrolases in maintaining cell integrity.

Results

PG is degraded in both spore types

To investigate if starvation-induced spores in M. xanthus fruiting bodies retain PG, we broke the fruiting bodies after 120 h of starvation and purified saculli from spores. For comparison, we also purified the saculli from vegetative cells and glycerol-induced spores. PG contents in all three cell types were analyzed using UPLC after muramidase digestion (see materials and Methods). The two spore types showed similar profiles of a discernible presence of muropeptides that resembled those found in vegetative cells, albeit in significantly reduced quantities (Fig. 1). Importantly, the most abundant muropeptides identified in the vegetative cells, MurNAc-tetrapeptide monomer (M4) and MurNAc-tetrapeptide dimer (D44), were nearly absent in both spore types (Fig. 1). Such residual muropeptides in both spore types are inadequate for forming continuous PG layers (Bui et al., 2009). This observation indicates that spore development in both pathways involves the breakdown of the vegetative PG cell wall.

Mature M. xanthus spores induced by either glycerol or starvation do not contain significant muropeptides.

UPLC muropeptides profiles of vegetative cells, glycerol– and starvation-induced spores indicate that the major muropeptides species, M4 and D44, in vegetative cells are diminished in both spore types, while spores are enriched in anhydro-muropeptides (Anh). The characteristic peaks are labeled as follows: M, monomeric muropeptide (uncrosslinked); D, dimeric muropeptide (crosslink connecting two muropeptides), T, trimeric muropeptide (crosslink connecting three muropeptides). Numbers refer to the status of the peptide side chain (3, tripeptide; 4, tetrapeptide). Red characters mark the muropeptides only detected in spores.

Despite the overall decline in muropeptides, anhydro-muropeptides markedly increased in both spore types, comprising over 90% of the total muropeptides (Fig. 1). Especially, anhydro-MurNAc-tetrapeptide trimer (T444Anh) that was only detected in trace amounts in vegetative cells, became a prominent muropeptide in both spore types (Fig. 1). Moreover, new anhydro-muropeptides, including anhydro, anhydro-MurNAc-tetrapeptide dimer (D44Anh,Anh), anhydro, anhydro-MurNAc-tetrapeptide trimer (T444Anh,Anh), and an anhydrodimer without a NAG (D44Anh(-NAG)), were only present in spores (Fig. 1). Because anhydro-muropeptides are the signature products of LTGs (Dik et al., 2017, Williams et al., 2018), their abundance in spores indicates that certain LTGs must play essential roles in M. xanthus sporulation.

The two sporulation pathways require different LTGs

The genome of M. xanthus encodes 14 putative LTGs (Aramayo & Nan, 2022, Ramirez Carbo et al., 2024). We imaged the cells of the 14 knockout mutants (Ramirez Carbo et al., 2024) at different time points after glycerol induction and used their length/width (L/W) ratios to monitor potential defects in the sporulation process. Wild-type cells initiated sporulation within 1 h of glycerol induction, which was evidenced by the increase of cell width and decrease of cell length (Fig. 2A). Consistent with the change in cell morphology, their L/W ratios decreased continuously and stabilized after 2 h of induction, when sporulation completed (Fig. 2B). In contrast, the cells that carry the deletion of open reading frame (ORF) K1515_20820 (MXAN_RS16290) (Aramayo & Nan, 2022) were able to shorten cell length slightly but retained rod shape after prolonged induction.

Glycerol-induced sporulation is regulated by both LtgA and LtgB, while starvation-induced sporulation only requires LtgB.

A) Bright field images of cells at different time points after glycerol-induction. Black arrows point to lysing cells. B) Quantitative analysis of glycerol-induced sporulation using the length/width ratio (L/W) of cells. Whiskers indicate the 25th – 75th percentiles and red dots the median. The total number of cells analyzed is shown on top of each plot. C) Only LtgB is required for fruiting body formation on starvation agar. D) LtgA and LtgB are homologous to E. coli MltE. The asterisk marks the conserved active site.

Surprisingly, another mutant that carries the deletion of ORF K1515_17460 (MXAN_RS19615) started to lose rod shape immediately after glycerol induction and became spherical within 1 h (Fig. 2A, 2B). However, different from the wild-type spores that appeared dark and heterogeneous under differential interference contrast (DIC) microscopy, the spheres formed by this mutant appeared bright and homogenous, similar to the pseudospores from the ΔaglQS mutant that lack the motor for depositing spore coat polysaccharides onto cell surfaces (Zhang et al., 2020) (Fig. 2A). To test whether these spheres are real spores, we subjected them to sonication and quantified their survival rate using a Helber bacterial counting chamber. After sonication, while 91 ± 6% (calculated from three independent experiments, n > 1,000, same below) of the wild-type spores appeared intact, only 3 ± 1% of the spheres produced by the K1515_17460 deletion mutant remained. Thus, this mutant indeed formed pseudospores that lacked the resistance against sonication. As the products of both K1515_20820 and K1515_17460 show homology to MltE, an LTG in Escherichia coli (encoded by emtA, Fig. 2D), we named them ltgA and ltgB, respectively. Both LtgA and LtgB are required for forming resistant spores via the rapid sporulation pathway, albeit playing opposite roles in the rod-to-sphere transition.

To identify the LTGs that are required for slow sporulation, we grew the 14 knockout strains in rich liquid media and spotted cells on solid starvation (CF) media. After 96 h of incubation, when wild-type cells formed dark fruiting bodies on the agar surface, the ΔltgB strain only formed flat aggregations (Fig. 2C). To test if ΔltgB cells could generate starvation-induced spores without forming fruiting bodies, we scraped such cell aggregations from the agar surface, suspended cells in rich liquid media, subjected them to sonication, then plated them on solid rich media. These cells failed to form colonies after five days of incubation, indicating that LtgB is essential for forming starvation-induced spores. In contrast, deleting ltgA did not affect fruiting body formation (Fig. 2C). Thus, the slow sporulation pathway only requires LtgB.

LtgA and LtgB play distinct roles in different sporulation pathways

Do LtgA and LtgB degrade PG during sporulation? To answer this question, we purified cell sacculi and imaged PG using immunofluorescence and an anti-PG serum (de Pedro et al., 1997). The sacculi of vegetative cells from the wild-type, ΔltgA, and ΔltgB strains were not visible under DIC microscope, likely due to their flattened structures minimizing light interference. However, PG from all three strains was readily detected in the fluorescence channel (Fig. 3A). After 6 h of glycerol-induction, the sacculi of both the wild-type and ΔltgA cells remained visible under DIC microscope (Fig. 3A), likely due to the deposition of spore coat polysaccharides that sustained unflattened cell structures. After the sacculus purification process, only background PG signals were detected in glycerol-induced wild-type spores (Fig. 3A). In contrast, the ΔltgA cells retained PG in substantial quantities (Fig. 3A). Sacculi from the ΔltgB pseudospores only contained a small amount of PG (Fig. 3A). Thus, compared to LtgA, the enzymatic activity of LtgB plays a more limited role in PG degradation during rapid sporulation. Different from the induced wild-type and ΔltgA cells, sacculi from the ΔltgB pseudospores were undetectable under DIC microscopy (Fig. 3A), reflecting the absence of rigid polysaccharide coats. Consequently, the residual PG in these pseudospores lost integrity during purification, with many sacculi displaying irregular shapes and ruptures in the fluorescence channel (Fig. 3A). Collectively, LtgB appears to be a pace-keeper that prevents abrupt PG degradation and thus allows sporulating cells to assemble polysaccharide coats during rapid sporulation.

LtgA and LtgB play distinct roles in the two sporulation pathways.

A) While LtgA is required for PG degradation during glycerol-induced sporulation, LtgB is the major LTG for forming starvation-induced spores. PG was detected using an anti-PG serum and a fluorescence-conjugated secondary antibody. PG sacculi were purified from cells after 6 h and 120 h of glycerol-induced and starvation-induced sporulation, respectively. Flattened sacculi are not visible under bright field microscopy. White arrows point to the spherical cells that show disintegrated PG. BF, bright field. B) The overexpression of LtgA, but not LtgB, collapses rod-shape in vegetative cells. C) Purified LtgA and LtgB solubilize dye-labeled PG sacculi at different rates. Lysozyme and buffer serve as the positive and negative controls, respectively. Absorption at 595 nm was measured after 18 h incubation at 25 ⁰C. Data are presented as mean values ± SD from three technical replicates. The inset shows purified LtgA and LtgB in a Coomassie stained gel. Scale bars, 5 μm.

To determine the roles of LtgA and LtgB in slow sporulation, we first scraped the wild-type and ΔltgA fruiting bodies from CF agar surface (Fig. 2C), dispersed spores by sonication, purified their sacculi, and visualized PG using immunofluorescence. Spherical spores from both the wild-type and ΔltgA cells were visible under DIC microscope, indicating that they retained unflattened shapes. These fruiting body spores lacked PG-specific fluorescence (Fig. 3A), likely due to both the absence of PG (Fig. 1) and the thickened polysaccharide coats (Voelz & Dworkin, 1962) preventing PG antibody access. Second, we investigated if ΔltgB cells degraded PG during starvation-induced sporulation. Because the flat aggregates formed by ΔltgB cells on CF agar did not contain mature spores, we scrapped cell aggregates from agar surface and purified their sacculi without sonication. Instead of forming spores, most of the ΔltgB cells were still rod-shaped, indistinguishable from vegetative ones (Fig. 3A). These observations indicate that LtgB is the only essential LTG for PG degradation during slow sporulation.

The two sporulation pathways vary greatly in duration: glycerol-induced rapid sporulation is completed within two hours, while starvation-induced slow sporulation unfolds over several days. Based on this distinction, we hypothesized that LtgA and LtgB could degrade PG at different rates. We used a vanillate-inducible promoter (Iniesta et al., 2012) to overexpress LtgA as a merodiploid. Induced by 200 µM vanillate, cells showed heterogeneous morphology, reflecting the variation in LtgA production. Many cells lost rod shape even in the absence of glycerol, indicating that these cells over-degraded their PG by LtgA (Fig. 2A, 3B). In contrast, the overexpression of LtgB using the same method did not affect the morphology of vegetative cells (Fig. 3B).

To further investigate the activities of these LTGs, we expressed the periplasmic domains of wild-type LtgA (amino acids 21 – 243) and LtgB (amino acids 26 – 710) in E. coli (Fig. 3C). We purified PG from wild-type M. xanthus cells, labeled it with Remazol brilliant blue (RBB), and tested if the purified AgmT variants hydrolyze labeled PG in vitro and release the dye (Jorgenson et al., 2014, Ramirez Carbo et al., 2024, Uehara et al., 2010). Wild-type LtgA solubilized dye-labeled PG, which absorbed light at 595 nm, demonstrating stronger hydrolytic activity than lysozyme, an enzyme that specifically cleaves β-1,4-glycosidic bonds in PG. In contrast, PG incubated with LtgB only showed minimum release of the dye, indicating slow PG hydrolysis (Fig. 3C). While we cannot rule out the possibility that our purification and reaction conditions were suboptimal for LtgB, both the in vitro RBB assay and the in vivo phenotype resulting from LtgB overexpression support our hypothesis that LtgA degrades PG at significantly higher rates than LtgB.

LtgB regulates LtgA during glycerol-induced sporulation

In line with its function in PG degradation, ltgA transcription increases approximately twofold during rapid sporulation but remains unchanged in slow sporulation. In contrast, ltgB shows a slight increase in expression during slow sporulation but remains unchanged in rapid sporulation (Muller et al., 2010, Munoz-Dorado et al., 2019). However, transcriptomic data alone do not account for the opposing roles of LtgA and LtgB in the rod-to-sphere transition during rapid sporulation. The resemblance in morphology between glycerol-induced ΔltgB cells and uninduced cells overproducing LtgA suggests that the slower-acting LtgB may offset the rapid activity of LtgA during sporulation, allowing sufficient time for spore coat polysaccharide deposition.

To test if LtgB regulates LtgA, we constructed a ΔltgA ΔltgB double deletion strain. Cells from this strain shortened their length slightly but failed to abolish rod shape after prolonged glycerol induction, phenocopying the sporulation defect of the ΔltgA strain (Fig. 2A, 2B). Hence, LtgB is a regulator upstream of LtgA.

LtgB exhibits a response to glycerol induction earlier than LtgA

Biochemical reactions on PG are unique for the stark size difference between the enzymes and their substrates. While the enzymes are in nanometer scales, their substrates, the PG sacculi, span several micrometers. Under microscope, PG remains stationary but PG-related enzymes are free to move. Even for M. xanthus that moves on surfaces, PG-related enzymes move at least two orders of magnitude faster than cell/PG (Zhang et al., 2023b, Ramirez Carbo et al., 2024). Thus, when diffusive enzymes bind to PG, their mobility decreases (Lee et al., 2016, Zhang et al., 2023b). By tracking single fluorescently-labeled enzyme particles, we can approximate their PG-binding in different physiological conditions and genetic backgrounds (Ramirez Carbo et al., 2024, Zhang et al., 2023b). We individually expressed photo-activatable mCherry (PAmCherry)-labeled LtgA and LtgB using their native loci and promoters. Both labeled LTGs accumulated as full-length proteins (Fig. S1A) and the PAmCherry tags did not affect the formation of either glycerol– or starvation-induced spores, indicating that these fusion proteins were fully functional (Fig. S1B, C). Consistent to the transcriptomic data (Muller et al., 2010, Munoz-Dorado et al., 2019), vegetative cells produce significantly less LtgA than LtgB (Fig. S1A).

We used a 405-nm excitation laser (0.3 kW/cm2, 0.1 s) to activate the fluorescence of a few labeled LTG particles randomly in each cell and quantified their localization using a 561-nm laser at 10 Hz using single particle tracking photo-activated localization microscopy (sptPALM, see Materials and Methods) (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013). As free PAmCherry particles diffuse extremely fast in the cytoplasm, entering and exiting the focal plane frequently, they usually appear as blurry objects that cannot be followed at 10 Hz close to the cell surface (Fu et al., 2018, Zhang et al., 2023b). For this reason, the noise from free PAmCherry due to potential protein degradation was negligible.

Single-particles of PAmCherry-labeled LtgA and LtgB can be categorized into two populations, immobile and mobile. The immobile particles remained within a single pixel (160 nm × 160 nm) before photobleach, and the mobile ones displayed typical diffusion (Fig. 4A). For the particles that switched between mobile and immobile states, our algorithm categorized them as mobile and calculated their diffusion coefficients (D) from their entire trajectories that contained both mobile and immobile segments. Hence, binding to PG not only increases the immobile population of the enzyme particles but also decreases the D of the mobile particles. In vegetative cells where large-scale PG degradation does not occur, 23.9% (n = 1175) and 18.2% (n = 824) of LtgA and LtgB particles were immobile, respectively (Fig. 4B). D values of the mobile population are 2.62 × 10-2 ± 2.0 × 10-3 µm2/s (n = 894) for LtgA-PAmCherry and 2.80 × 10-2 ± 2.50 × 10-3 µm2/s (n = 674) for LtgB-PAmCherry (Fig. 4C).

LtgB regulates the PG-binding of LtgA during glycerol-induced sporulation.

A) LtgB exhibits a response to glycerol induction earlier than LtgA. Representative trajectories of LtgA and LtgB before (uninduced) and after (1 min and 30 min) glycerol induction. The overall distribution of both LTGs is displayed using the composite of 100 consecutive frames taken at 100-ms intervals. Single-particle trajectories of PAmCherry were generated from the same frames. Individual trajectories are distinguished by colors. Scale bars, 5 μm. B and C) the absence of LtgB reduces the diffusion of LtgA, which is reflected in the increase of immobile population (B) and the decrease in D (C), and these effects are especially prominent in the cells before (uninduced) and immediately after (1 min) glycerol induction. For each protein and condition, particles were identified from at least 100 cells and three independent experiments. The total number of particles analyzed is shown on top of each plot. Error bars were the standard derivation of 1,000 bootstrap samples and * indicates a significant difference of > 0.005.

We then determined how the dynamics of both LTGs respond to glycerol induction in the rapid sporulation pathway. Immediately (1 min) after adding glycerol, the dynamics of LtgA remained little changed, when 21.0% (n = 928) of particles were immobile and the D of the mobile ones was 2.53 × 10-2 ± 0.19 × 10-3 µm2/s (n = 733) (Fig. 4B, 4C). In contrast, the mobility of LtgB decreased significantly, with the immobile population increased to 28.0% (n = 1359) and D of the mobile population decreased to 2.17 × 10-2 ± 2.75 × 10-3 µm2/s (n = 978) (Fig. 4B, 4C). Therefore, these results indicate that upon glycerol induction, LtgB rapidly enhances its binding to PG.

We then chose 30 min after glycerol induction as a time point for rapid PG degradation, which is reflected by the dramatic decrease of L/W during the first hour of sporulation (Fig. 2A, 2B). Compared to its inert response 1 min after glycerol induction, the mobility of LtgA particles decreased significantly at 30 min, when the immobile population increased to 31.5% (n = 1163) and D of the mobile population decreased to 1.80 × 10-2 ± 2.12 × 10-3 µm2/s (n = 797) (Fig. 4B, 4C). The reduced mobility of LtgA aligns with its function as the primary LTG in glycerol-induced sporulation. Strikingly different from LtgA, as sporulation advanced to the 30-min mark, LtgB’s mobility returned to its pre-induction level, with the immobile population decreased to 10.2% (n = 1549) and D of the mobile population increased to 2.87 × 10-2 ± 2.76 × 10-3 µm2/s (n = 1391) (Fig. 4B, 4C). Therefore, the PG-binding of LtgB shows a negative correlation with PG degradation. Taken together, LtgA and LtgB display opposite responses to glycerol induction, in which LtgB rapidly binds to PG before yielding to LtgA, whose PG-binding is concurrent with the rod-to-sphere transition. The sequential PG-binding by these two LTGs controls the pace of PG-degradation during glycerol-induced sporulation.

LtgB blocks LtgA from binding PG in the early stage of glycerol-induced sporulation

Does LtgB’s early PG-binding suppress PG degradation by LtgA? To answer this question, we expressed LtgA-PAmCherry using the native ltgA locus and promoter in the ΔltgB background. In vegetative cells, the absence of LtgB significantly reduced the mobility of single LtgA-PAmCherry particles, suggesting that LtgB does affect LtgA’s binding to the PG (Fig. 4B, 4C). Similarly, we expressed LtgB-PAmCherry using the native ltgB locus and promoter in the ΔltgA background. In contrast, the mobility of single LtgB-PAmCherry particles only decreased slightly in the absence of LtgA (Fig. 4B, 4C). These results suggest that while LtgB significantly reduces LtgA’s binding to PG, probably due to its higher expression level, LtgA has only a modest impact on LtgB’s ability to bind PG.

Strikingly different from its slow response to glycerol in wild-type cells, LtgA increased its PG-binding immediately (1 min) after glycerol induction in the ΔltgB background, confirming that LtgB restricts LtgA access to the PG in the early stage of sporulation (Fig. 4B, 4C). In contrast, at 30 min of induction, a time point when LtgB dissociated from PG, its absence no longer affected LtgA’s PG-binding (Fig. 4B, 4C). Similar to the observation in wild-type cells, the lack of LtgA did not affect the PG-binding of LtgB at either 1 min or 30 min after glycerol induction (Fig. 4B, 4C). In summary, LtgB, a slow LTG, regulates the progression of rapid sporulation by preventing LtgA, a fast LTG, from excessively degrading PG, thereby ensuring that sporulating cells can deposit polysaccharide spore coats before PG disintegration occurs.

Upregulated PG synthesis negatively affects PG degradation

Since LtgA is less mobile in vegetative ΔltgB cells, i.e. strongly binds to PG (Fig. 4B, C), why do these cells retain their rod shape rather than losing it spontaneously before glycerol induction? In addition to PG-binding, LtgA may require additional stimuli to initiate PG degradation. We hypothesized that, similar to certain antibiotics targeting PG synthases, which induce wild-type cells to degrade PG and form pseudospores in rich liquid media (Zhang et al., 2023b, O’Connor & Zusman, 1997), glycerol may alter cellular metabolism, leading to reduced PG synthesis. If this is the case, cells with upregulated PG synthesis should remain rods even in the presence of glycerol.

To test our hypothesis, we used the vanillate-inducible promoter (Iniesta et al., 2012) to overexpress murA as a merodiploid in the wild-type background. Because MurA catalyzes the first committed step of PG synthesis that produces UDP-MurNAc (Egan et al., 2020, Rohs & Bernhardt, 2021), elevated MurA levels are expected to channel more cellular resources toward PG synthesis. Overproduction of MurA in the presence of 200 µM vanillate resulted in a heterogeneous cell population, with normal cells coexisting alongside elongated ones (Fig. 2A, B). This heterogeneity may reflect the heterogeneous induction of murA across the whole population. Overall, excessive MurA increased the average length of vegetative cells by 19.6%. Cells overproducing MurA also displayed heterogeneity during glycerol-induced sporulation. While some cells transitioned into spheres, many retained their rod shape even after 6 h of induction (Fig. 2A, B). Notably, sporadic lysis of rod-shaped cells began 2 h post-induction and became increasingly frequent with continued incubation (Fig. 2A). Potentially, elevated muropeptide production may lead to the accumulation of toxic intermediates that the relatively insufficient LTGs failed to degrade (Weaver et al., 2022). Nevertheless, the capacity of MurA overexpression to enable certain cells to circumvent glycerol-induced sporulation implies that PG production acts as a regulatory cue for PG degradation.

To test if excessive MurA can also reduce PG degradation during slow sporulation, we induced MurA overexpression on solid CF agar containing 100 µM vanillate. After 96 h of incubation, these cells developed fruiting bodies that were noticeably larger and flatter compared to those of the wild-type strain (Fig. 2C). To test if such fruiting bodies contained mature spores, we scraped them from the agar surface and tested the germination rate of spores on CYE agar after sonication. The fruiting bodies formed by cells overexpressing MurA yielded 21.3% of the colonies produced by an equivalent number of wild-type cells, indicating that upregulated PG synthesis also prevents slow sporulation. In contrast, cells grown without vanillate progressed normally through both sporulation pathways, showing no differences compared to wild-type cells (Fig. S2). Therefore, the sporulation of M. xanthus is sensitive to PG synthesis and increased PG synthesis balances PG degradation in both sporulation pathways.

Discussion

It is still unknown how many bacteria outside of the Firmicutes phylum form true spores. Nevertheless, our findings elucidate a new mechanism of non-firmicute bacterial sporulation that depends on PG degradation. This mechanism presents a significant challenge for sporulating cells: preserving their structural integrity while simultaneously dismantling the primary framework that upholds it. Our findings revealed that besides regulating the production of LTGs, cells could control the pace of PG degradation in a rapid sporulation pathway through the regulation between two LTGs and hence allow the resistant polysaccharide coats to assemble and mature.

Studying the regulation between two LTGs is especially challenging both in vivo and in vitro for a few reasons. First, it is difficult to monitor their instantaneous activities in vivo. Secondly, muropeptide analysis may not distinguish their roles because their products bear the same signature, anhydro-MurNAc. Thirdly, some enzymes require special substrates or activators, which would be difficult to study in vitro. To overcome these technical hurdles, we leveraged single-particle mobility to quantify the PG-binding of these enzymes. The simultaneous occurrence of reduced LtgA mobility and PG degradation during glycerol-induced sporulation indicates that the molecular dynamics of LtgA accurately mirrors its enzymatic activity. Such correlation between decreased particle mobility and increased enzymatic activity applies to many other PG-related enzymes, including multiple PG polymerases in E. coli and the endopeptidase DacB in M. xanthus (Lee et al., 2016, Zhang et al., 2023b, Yang et al., 2021). Because the size difference between PG and its related enzymes is a conserved feature in all bacteria, our method using single-particle tracking to quantify PG-binding can be applied in many organisms.

We discovered that LtgB, a slower-acting LTG, limits the access of the faster LtgA to the PG during the early phase of rapid sporulation, thereby regulating the rate of its degradation. Then how does LtgA breach the blockage of LtgB and gain access to PG as sporulation progresses? Because the overexpression of LtgA is sufficient to damage PG in uninduced cells (Fig. 3B), we propose that the relative abundance of these two LTGs plays a critical role in determining the fate during fast sporulation. While LtgB nearly saturates PG binding sites in vegetative cells where few LtgA are produced, LtgA— upregulated after glycerol induction (Muller et al., 2010) —gradually outcompetes LtgB and thereby causes LtgB to disassociate from PG.

As increased PG synthesis enables many cells to bypass both sporulation pathways, it is reasonable to propose that diminished PG synthesis initiates PG degradation by LTGs. While functional coordination between PG synthesis and hydrolysis has been speculated for over 50 years (Koch, 1985, Koch, 1990), its mechanisms only came to light recently. Some endopeptidases are proposed to serve as the space-makers for PG synthases (Dorr et al., 2013, Singh et al., 2012). In M. xanthus, we reported that moenomycin, an antibiotic that specifically binds to the GTase domains of class A penicillin-binding proteins (aPBPs) (Sung et al., 2009, Lovering et al., 2007), specifically activates a PG endopeptidase DacB through PBP1a2, an aPBP (Zhang et al., 2023b). Because moenomycin mimics a growing glycan strand in the GTase domains of aPBPs, it actually locks aPBPs in their glycan-charged conformations (Sung et al., 2009, Lovering et al., 2007). Thus, similar to the moenomycin-bound form, glycan-charged PBP1a2 in physiological conditions recruits DacB to the PG assembly sites, which in turn, generates openings in the existing PG network as the crosslink sites for the TPase activity of PBP1a2. Such “make-before-break” mechanism (Koch & Doyle, 1985) could prevent unneeded hydrolysis and thus maintain PG integrity.

Compared to the well-characterized space-making functions of endopeptidases, the roles of LTGs, beyond PG recycling, remain ill-defined. Even less is known on their potential coordination with PG synthases (Weaver et al., 2023). Because LTGs from bacterial and phage origins are inhibited by L,D-crosslinks, they may functionally connect to PBPs that form D,D-crosslinks (Alvarez et al., 2024). Additionally, some LTGs, such as MltD and MltG in E. coli and MltG in Vibrio cholerae, prefer to degrade nascent, uncrosslinked glycan strands (Kaul et al., 2024, Weaver et al., 2022, Bohrhunter et al., 2021). Our recent report indicates that AgmT, an M. xanthus LTG homologous to E. coli MltG, detoxifies uncrosslinked glycan strands under antibiotic stress and modifies crosslinked PG scaffolds to attach a motility machinery to PG (Ramirez Carbo et al., 2024). Then how does reduced PG synthesis activate LtgA in M. xanthus? Although MltE, the closest homolog of LtgA in E. coli, degrades glycan strands regardless of their crosslinking status or peptide content, it shows a marked preference for uncrosslinked substrates (Dik et al., 2017, Fibriansah et al., 2012). Similar to AgmT and MltE, LtgA might also bind to both crosslinked and uncrosslinked glycan strands. Consistent with its high mobility, LtgA may coordinate with diffusive PG polymerases and cleave the uncrosslinked glycan strands. During rapid sporulation, as nascent glycan strands decline due to muropeptide depletion, LtgA increasingly binds to the crosslinked PG scaffold, which is evidenced by its reduced mobility. This binding intensifies further as PG degradation continues, establishing a positive feedback loop. In this case, LtgB serves as a brake that dampens this positive feedback, moderates PG degradation, and thus allows sporulating cells to maintain integrity by assembling polysaccharide coats. Such cross-regulation between PG hydrolases, which is currently under investigation, provides yet another mechanism by which bacteria dynamically maintain their PG cell walls.

Materials and methods

Bacterial strains and growth conditions

Vegetative M. xanthus cells were grown in liquid CYE medium (10 mM MOPS pH 7.6, 1% (w/v) Bacto™ casitone (BD Biosciences), 0.5% yeast extract and 8 mM MgSO4) at 32 °C, in 125-ml flasks with vigorous shaking, or on CYE plates that contains 1.5% agar. We used strain DZ2 as the wild-type M. xanthus strain (Campos & Zusman, 1975). Knock-out mutants were constructed by electroporating DZ2 cells with 4 µg of plasmid DNA. Transformed cells were plated on CYE plates supplemented with 100 mg/ml sodium kanamycin sulfate or 10 mg/ml tetracycline. The strains and plasmids used in this study are listed in Table 1.

Sporulation and spore purification

Cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2. Glycerol was added to 1 M to induce sporulation. The liquid culture was incubated at 32 °C with vigorous shaking. After 24 h, spores and cells were harvested by centrifugation (10 min, 10,000 g and 25 °C). To further purify the wild-type spores, the pellet was resuspended in 5 ml water, the remaining vegetative cells were eliminated by sonication (Cole Palmer 4710 ultrasonic homogenizer, 30% output, 10 cycles), and sonication-resistant spores were washed three times with water and collected by centrifugation (5 min, 10,000 g and 4 °C). The elimination of vegetative cells was confirmed by DIC microscopy.

For phenotypic assays on starvation-induced fruiting body formation, vegetative cells (10 ul), at a concentration of 4 × 109 colony-forming units (cfu) ml-1, were spotted on CF (0.015% Casitone, 0.2% sodium citrate, 0.1% sodium pyruvate, 0.02% (NH4)2SO4, 10 mM MOPS (pH 7.6), 8 mM MgSO4 and 1 mM KH2PO4) plates containing an agar concentration of 1.5%, incubated at 32°C. To purify fruiting body spores, cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2 and harvested by centrifugation (10 min, 10,000 g and 25 °C). The pellet was resuspended in 1 ml water and plated on two 150 mm CF plates containing 1.5% agar using a spreader. The plates were air-dried and incubated at 32 °C for 120 h. For the wild-type and ΔltgA strains, fruiting bodies were scraped from the plates, suspended in 2 ml water, and subjected to sonication (Cole Palmer 4710 ultrasonic homogenizer, 60% output, 20 cycles). Spores were wash three times in water and collected by centrifugation (5 min, 10,000 g and 4 °C). For the ΔltgB strains, the sonication step was omitted.

Peptidoglycan purification and UPLC analysis

For vegetative cell peptidoglycan analysis, samples were processed as previously described for Gram negative bacteria (Alvarez et al., 2016, Desmarais et al., 2013). In brief, M. xanthus cells were grown until mid-stationary phase and harvested by centrifugation (30 min, 8,000 g). Supernatant was discarded and the pellet was resuspended and boiled in 1x PBS with 5% SDS for 2 h. The saccules were repeatedly washed with MilliQ water by ultracentrifugation (150,000 g, 10 min, 20°C) to remove the remaining SDS. The samples were then treated with muramidase (100 µg/mL) for 16 hours at 37 °C. Muramidase digestion was stopped by boiling and coagulated proteins were removed by centrifugation (10 min, 15,000 g). The supernatants were first adjusted to pH 8.5-9.0 with sodium borate buffer and then sodium borohydride was added to a final concentration of 10 mg/mL. After reducing the samples at room temperature for 30 mins, the pH was adjusted to pH 3.5 with orthophosphoric acid.

UPLC analyses of muropeptides were performed on a Waters UPLC system (Waters Corporation, USA) equipped with an ACQUITY UPLC BEH C18 Column, 130Å, 1.7 µm, 2.1 mm X 150 mm (Waters, USA) and a dual wavelength absorbance detector. Elution of muropeptides was detected at 204 nm. Muropeptides were separated at 45 °C using a linear gradient from buffer A (formic acid 0.1% in water) to buffer B (formic acid 0.1% in acetonitrile) in a 25-min run, with a 0.30 mL/min flow. For vegetative cells analysis, relative total PG amounts were calculated by comparison of the total intensities of the chromatograms (total area) from three biological replicas normalized to the same initial biomass and extracted with the same volumes. Quantification of muropeptides was based on their relative abundances (relative area of the corresponding peak).

Imaging and data analysis

For all imaging experiments on isolated spores/cells, we spotted 5 μl of cells grown in liquid CYE medium to OD600 ∼1 on agar (1.5%) pads and imaged using a Andor iXon Ultra 897 EMCCD camera (effective pixel size 160 nm) on an inverted Nikon Eclipse-Ti™ microscope with a 100✕ 1.49 NA TIRF objective. Fruiting bodies were photographed after a 120-h incubation on CF agar using a Nikon SMZ1000 microscope and an OMAX A3590U digital camera. Cell morphology was imaged at indicated time points using DIC microscopy. The geometric aspect ratios (L/W) of spores/cells were determined using a custom algorithm written in MATLAB, which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b).

For sptPALM, M. xanthus cells were grown in CYE to 4 ×108 cfu/ml, spotted on 1.5% agar pads and subjected to highly inclined and laminated optical sheet (HILO) illumination (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013, Tokunaga et al., 2008). PAmCherry was activated using a 405-nm laser (0.3 kW/cm2), excited and imaged using a 561-nm laser (0.2 kW/cm2). Images were acquired at 10 Hz. For each sptPALM experiment, single PAmCherry particles were localized in at least 100 individual cells from three biological replicates. sptPALM data were analyzed using a MATLAB (MathWorks) script, which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b). Briefly, cells were identified using differential interference contrast images. Single PAmCherry particles inside cells were fit by a symmetric 2D Gaussian function, whose center was assumed to be the particle’s position (Fu et al., 2018). Particles that explored areas smaller than 160 nm × 160 nm (within one pixel) in 0.4 – 1.2 s were considered immotile (Fu et al., 2018, Zhang et al., 2023a, Zhang et al., 2023b). D of all the mobile particles was determined from a linear fit to the first four points of the MSD using a formula MSD = 4DΔt (Fu et al., 2018, Lee et al., 2016). Error bars were the standard derivation of 1,000 bootstrap samples using the published method (Morgenstein et al., 2015). Sample trajectories were generated using the TrackMate (Ershov et al., 2022) plugin in the ImageJ suite (https://imagej.net).

Immunofluorescence

Cell samples were resuspended and boiled in 1x PBS with 5% sodium dodecyl sulfate (SDS) for 2 h. The saccules were repeatedly washed with water by centrifugation (5 min, 15,000 g and 25 °C) to remove the remaining SDS. Remaining proteins were further removed by incubation with proteinase K (1 µg/ml) for 2 h at 37 °C. Microscope cover slides were prepared by coating with 100 µl 0.1% poly-L-lysine for 5 min at room temperature (RT), washing with ddH2O and air drying. 100 µl samples were spotted onto the slides and incubated for 20 min at RT. The slides were then washed three times with phosphate buffer saline (PBS; 81 mM Na2HPO4, 15 mM KH2PO4, 1.37 M NaCl, 27 mM KCl, pH 7.4), and then incubated with anti-PG primary antibody16 diluted 1:1000 in PBS at RT for 1 h. Slides were washed 6 times with PBS and then incubated with Alexa Fluor 488 conjugated goat-anti-rabbit secondary antibody (Invitrogen) at a 1:1000 dilution in PBS at RT for 1 h in the dark. Slides were washed 6 times with PBS. Each cover glass was applied to a supporting slide, fixed with nail polish, and stored at 4 °C, if necessary. Images were recorded using a 488-nm laser.

Protein expression and purification

DNA sequences encoding amino acids 21 – 243 of LtgA and 26 – 710 of LtgB were amplified by polymerase chain reaction (PCR) and inserted into the pET28a vector (Novagen) between the restriction sites of EcoRI and HindIII and used to transform E. coli strain BL21(DE3). Transformed cells were cultured in 20 ml LB (Luria-Bertani) broth at 37 °C overnight and used to inoculate 1 L LB medium supplemented with 1.0% glucose. Protein expression was induced by 0.1 mM IPTG (isopropyl-h-d-thiogalactopyranoside) when the culture reached an OD600 of 0.8. Cultivation was continued at 16 °C for 10 h before the cells were harvested by centrifugation at 6,000 × g for 20 min. Proteins were purified using a NGC™ Chromatography System (BIO-RAD) and 5-ml HisTrap™ columns (Cytiva) (Pogue et al., 2018, Nan et al., 2010). Purified proteins were concentrated using Amicon™ Ultra centrifugal filter units (Millipore Sigma) with a 10-kDa molecular weight cutoff and stored at –80 °C.

LTG activity (RBB) assay

PG was purified following the published method (Zhang et al., 2023b, Alvarez et al., 2016, Ramirez Carbo et al., 2024). In brief, M. xanthus cells were grown until mid-stationary phase and harvested by centrifugation (6,000 × g, 20 min, 25 °C). Supernatant was discarded and the pellet was resuspended and boiled in 1× PBS with 5% SDS for 2 h. SDS was removed by repetitive wash with water and centrifugation (21,000 × g, 10 min, 25 °C). Purified PG from 100 ml culture was suspended into 1 ml 1× PBS and stored at –20 °C. RBB labelling of PG was performed essentially as previously described (Uehara et al., 2010, Jorgenson et al., 2014). Purified sacculi were incubated with 20 mM RBB in 0.25 M NaOH overnight at 37 °C. Reactions were neutralized by adding equal volumes of 0.25 M HCl and RBB-labeled PG was collected by centrifugation at 21,000 × g for 15 min. Pellets were washed repeatedly with water until the supernatants became colorless. RBB-labelled sacculi were incubated with purified LtgA and LtgB (1 mg/ml) at 25 °C for 12 h. lysozyme (1 mg/ml) was used as a positive control. Dye release was quantified by the absorption at 595 nm from the supernatants after centrifugation (21,000 × g, 10 min, 25 °C).

eLife Assessment

This important study identifies and partially characterises two proteins optimised for coordinated peptidoglycan degradation during two spore morphogenesis programs in the bacterium Myxococcus xanthus. The evidence supporting the conclusions is solid; however, the description of the data is somewhat overstated. After some editing, the paper will be of interest to those studying peptidoglycan synthesis and reorganisation, which is a central aspect of microbial cell biology.

Reviewer #1 (Public review):

Summary:

Ramirez Carbo et al. use the powerful M. xanthus spore morphogenesis model to address fundamental mechanisms in coordinated peptidoglycan remodeling and degradation. As peptidoglycan is an essential macromolecule and difficult to study in vivo, the authors use indirect but important methodology. The authors first identify two lytic transglycosylase (Ltg) enzymes necessary for spore morphogenesis using mutant phenotypic studies. They characterize these mutants for their role in coordinating spore morphogenesis induced either in fruiting bodies (starvation-dependent) or in liquid-rich media conditions (chemical-dependent). They conclude from these phenotypic and epistatic analyses that LtgA is necessary for morphogenesis during chemical-induced sporulation, and LtgB appears to be necessary to coordinate LtgA activity by interfering with LtgA function. Under starvation-induced sporulation, the absence of LtgB interferes with the building of fruiting bodies. LtgA does not appear to play a primary role in promoting aggregation into fruiting bodies, nor in degradation of peptidoglycan as assayed by loss of signal in anti-PG immunofluorescence. The authors demonstrate that the purified periplasmic domain of LtgA is highly active in degrading purified PG sacculi in vitro, while that of LtgB is highly reduced (relative to LtgA or lysozyme). The authors use photoactivated mCherry Lyt fusions and PALM to track the fusion protein mobility, which they state correlates with activity as immobilization results from PG binding. They demonstrate that in vegetative cells, a greater proportion of LtgA-PAmCh is more immobile (more active) than LtgB-PAmCh, but that directly after chemical-induction of sporulation, LtgB-PAmCh becomes more immobile (active). These analyses in the partner mutant backgrounds suggest that LtgA-PAmCh is more immobile (less active) in the absence of LtgB, but the reverse is not observed. Finally, the authors demonstrate that overexpression of LtgA in vegetative conditions leads to cell rounding, likely because of uncontrolled PG degradation, while overexpression of LtgB displays no phenotype.

Strengths:

This paper capitalizes on a novel spore morphogenesis mechanism to define proteins and mechanisms involved in peptidoglycan reorganization. The authors use the powerful PALM microscopy technique to assess Ltg activity in vivo by assaying for immobility as a proxy for PG binding. The authors elucidate a novel mechanism by which two Ltg's function together- with one (LtgB) seeming to regulate the activity of the other (the primary Ltg).

Despite some weaknesses, there is no question that this study provides important insight into mechanisms of peptidoglycan remodeling- a difficult but highly impactful area of study with implications for the development of novel therapeutics and the discovery of mechanisms of fundamental bacterial physiology.

Weaknesses:

In many places, the authors do not adequately justify interpretations of their assays, leading to some apparently unjustified conclusions. Many of these are minor and may just require citations to demonstrate that the interpretations are justified by previous studies (detailed in recommendations below), but two bigger concerns are as follows:

(1) It is not clear how the muropeptides listed in Figure 1 were assigned, and it is missing in the methods. In the sporulating conditions, the spectra look like combinations of multiple peaks, and the data, as stated, is not convincing to the non-specialist eye.

(2) The observation that the lytB mutant prevents appropriate aggregation into fruiting bodies does not allow the interpretation that the absence of LytB prevents PG morphogenesis in the starvation-induced sporulation pathway, per se. It is more likely that in the lytB mutant, the morphogenesis program is not even triggered. This is because signaling proteins and regulators (specifically, C-signal accumulation/activated FruA), which are dependent on increased cell-cell signaling in the fruiting body, do not accumulate appropriately in shallow aggregates. C-signal/FruA are necessary to trigger the sporulation program in FBs. BTW: A hypothesis to explain the indirect effect of ltgB absence on aggregation could be that UDP-precursors are not regulated appropriately (unregulated LtyA??), so polysaccharides necessary for motility are not properly produced.

Along these lines, fruiting body formation does not equal sporulation, and even "darkened" fruiting bodies can be misleading, as some mutants form polysaccharide-rich fruiting bodies (that appear dark under certain light conditions in the stereomicroscope) but do not sporulate efficiently. The wording in the text suggests that the authors assume that sporulation levels are normal because fruiting bodies are produced (see specific comments for details).

(3) The authors repeatedly state that production of spore coat polysaccharides likely affects the PG IP staining (see below), but this is not well justified. A citation is needed if this has already been directly shown, or the language needs to be softened.

(4) Better justification for the immobility of Lyt proteins in vivo as an assay for activity may be required. If this is well known in the field, it should be explicitly stated. The authors address this better in the discussion - but still state it is a correlation.

Reviewer #2 (Public review):

Summary:

The authors' initial goal was to demonstrate loss of PG during the slow sporulation process of Myxococcus xanthus, with examination of the PG degradation products in order to implicate possible enzymes involved. Upon finding a predominance of LGT products, they examined sporulation in strains lacking each of the 14 candidate LGTs encoded in the genome, leading to the identification of two sporulation-linked LGTs. An extensive characterization of the roles played by these LGTs. One LGT is responsible for the slow sporulation PG degradation, while another is required for the rapid sporulation process. Interestingly, the "slow" LGT seems to provide an important regulatory brake on the rapid enzyme. Single-molecule fluorescent tracking of these enzymes was used to develop a model for their interaction with PG that mimics their observed activity. The rate of PG synthesis activity was also shown to impact the rate of PG degradation, suggesting potential interplay between the synthetic and degradative enzymes.

Strengths:

The genetic analysis to identify sporulation-linked LGTs and their effects on growth, sporulation, and spore properties was well done and productive. The fluorescence microscopy to track LGT mobility, presumably tied to activity, produced a convincing argument about the mechanism of regulation of one LGT by another.

Weaknesses:

While the impact of LGTs on sporulation was clearly demonstrated, the PG analysis that resulted from the study of LGTs raised some important unanswered questions. The analyses suggest that the PG is degraded to quite small fragments, which would normally be lost during the purification of PG. How these small fragments were thus detected is unclear, and this suggests a more complex story concerning PG metabolism during sporulation. An anti-PG antibody is used to quantify PG in the spores, but it is not made clear what the specificity of this antibody is, and thus whether it would recognize the LGT-altered PG of the spore. The authors suggest a "new mechanism of sporulation" when they have actually simply identified an important factor (PG degradation by LGTs) within a complex "process of sporulation".

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Ramirez Carbo et al. use the powerful M. xanthus spore morphogenesis model to address fundamental mechanisms in coordinated peptidoglycan remodeling and degradation. As peptidoglycan is an essential macromolecule and difficult to study in vivo, the authors use indirect but important methodology. The authors first identify two lytic transglycosylase (Ltg) enzymes necessary for spore morphogenesis using mutant phenotypic studies. They characterize these mutants for their role in coordinating spore morphogenesis induced either in fruiting bodies (starvation-dependent) or in liquid-rich media conditions (chemical-dependent). They conclude from these phenotypic and epistatic analyses that LtgA is necessary for morphogenesis during chemical-induced sporulation, and LtgB appears to be necessary to coordinate LtgA activity by interfering with LtgA function. Under starvation-induced sporulation, the absence of LtgB interferes with the building of fruiting bodies. LtgA does not appear to play a primary role in promoting aggregation into fruiting bodies, nor in degradation of peptidoglycan as assayed by loss of signal in anti-PG immunofluorescence. The authors demonstrate that the purified periplasmic domain of LtgA is highly active in degrading purified PG sacculi in vitro, while that of LtgB is highly reduced (relative to LtgA or lysozyme). The authors use photoactivated mCherry Lyt fusions and PALM to track the fusion protein mobility, which they state correlates with activity as immobilization results from PG binding. They demonstrate that in vegetative cells, a greater proportion of LtgA-PAmCh is more immobile (more active) than LtgB-PAmCh, but that directly after chemical-induction of sporulation, LtgB-PAmCh becomes more immobile (active). These analyses in the partner mutant backgrounds suggest that LtgA-PAmCh is more immobile (less active) in the absence of LtgB, but the reverse is not observed. Finally, the authors demonstrate that overexpression of LtgA in vegetative conditions leads to cell rounding, likely because of uncontrolled PG degradation, while overexpression of LtgB displays no phenotype.

Strengths:

This paper capitalizes on a novel spore morphogenesis mechanism to define proteins and mechanisms involved in peptidoglycan reorganization. The authors use the powerful PALM microscopy technique to assess Ltg activity in vivo by assaying for immobility as a proxy for PG binding. The authors elucidate a novel mechanism by which two Ltg's function together- with one (LtgB) seeming to regulate the activity of the other (the primary Ltg).

Despite some weaknesses, there is no question that this study provides important insight into mechanisms of peptidoglycan remodeling- a difficult but highly impactful area of study with implications for the development of novel therapeutics and the discovery of mechanisms of fundamental bacterial physiology.

Weaknesses:

In many places, the authors do not adequately justify interpretations of their assays, leading to some apparently unjustified conclusions. Many of these are minor and may just require citations to demonstrate that the interpretations are justified by previous studies (detailed in recommendations below), but two bigger concerns are as follows:

(1) It is not clear how the muropeptides listed in Figure 1 were assigned, and it is missing in the methods. In the sporulating conditions, the spectra look like combinations of multiple peaks, and the data, as stated, is not convincing to the non-specialist eye.

We thank the reviewer for raising this point. We've expanded the Methods section to give a fuller account of how muropeptides were identified. In particular, we now describe the chromatographic separation and the assignment process, which relies on comparison with published data, fragmentation patterns, retention times, and accurate mass values. We acknowledge that the chromatograms show several peaks; nevertheless, only those muropeptides that we could clearly identify by MS/MS analysis were annotated in Figure 1. This clarification is now made explicit in the figure legend in the revised version of the manuscript.

(2) The observation that the lytB mutant prevents appropriate aggregation into fruiting bodies does not allow the interpretation that the absence of LtgB prevents PG morphogenesis in the starvation-induced sporulation pathway, per se. It is more likely that in the LtgB mutant, the morphogenesis program is not even triggered. This is because signaling proteins and regulators (specifically, C-signal accumulation/activated FruA), which are dependent on increased cell-cell signaling in the fruiting body, do not accumulate appropriately in shallow aggregates. C-signal/FruA are necessary to trigger the sporulation program in FBs. BTW: A hypothesis to explain the indirect effect of ltgB absence on aggregation could be that UDP-precursors are not regulated appropriately (unregulated LtyA (LtgA [sic])??), so polysaccharides necessary for motility are not properly produced.

Along these lines, fruiting body formation does not equal sporulation, and even "darkened" fruiting bodies can be misleading, as some mutants form polysacchariderich fruiting bodies (that appear dark under certain light conditions in the stereomicroscope) but do not sporulate efficiently. The wording in the text suggests that the authors assume that sporulation levels are normal because fruiting bodies are produced (see specific comments for details).

We deeply appreciate this question. Seeking the answer, we repeated the fruiting body assay and found that both the ΔltgA and ΔltgB mutants formed dark aggregates that were comparable to wild-type fruiting bodies. However, these “fruiting body-like” aggregates did not contain sonication-resistant spores. Thus, regardless of the signals, either glycerol or starvation, sporulation requires both LtgA and LtgB. We have corrected the mistakes in the first submission.

(3) The authors repeatedly state that production of spore coat polysaccharides likely affects the PG IP staining (see below), but this is not well justified. A citation is needed if this has already been directly shown, or the language needs to be softened.

We agree with the reviewer. We have softened our language as “However, we cannot exclude the possibility that the polysaccharide spore coats (Voelz & Dworkin, 1962) hinder antibody access to PG.”

(4) Better justification for the immobility of Ltg proteins in vivo as an assay for activity may be required. If this is well known in the field, it should be explicitly stated. The authors address this better in the discussion - but still state it is a correlation.

We elaborated the justification, “Thus, when diffusive enzymes bind to PG, their mobility decreases (Lee et al., 2016; Zhang et al., 2023). For instance, DacB, another PG hydrolase, reduces its single-particle mobility in the conditions where its activity is activated (Zhang et al., 2023). By tracking single fluorescently-labeled enzyme particles, we can approximate their PG-binding in different physiological conditions and genetic backgrounds (Ramirez Carbo et al., 2024, Zhang et al., 2023, Ramírez Carbó & Nan, 2026).”

We further discussed the correlation in discussion, “The simultaneous occurrence of reduced LtgA mobility and PG degradation during glycerol-induced sporulation indicates that the molecular dynamics of LtgA accurately mirrors its enzymatic activity. Such correlation between decreased particle mobility and increased enzymatic activity applies to many other PG-related enzymes, including multiple PG polymerases in E. coli and the endopeptidase DacB in M. xanthus (Lee et al., 2016, Zhang et al., 2023, Yang et al., 2021).”

Reviewer #2 (Public review):

Summary:

The authors' initial goal was to demonstrate loss of PG during the slow sporulation process of Myxococcus xanthus, with examination of the PG degradation products in order to implicate possible enzymes involved. Upon finding a predominance of LGT products, they examined sporulation in strains lacking each of the 14 candidate LTGs encoded in the genome, leading to the identification of two sporulation-linked LTGs. An extensive characterization of the roles played by these LTGs. One LTG is responsible for the slow sporulation PG degradation, while another is required for the rapid sporulation process. Interestingly, the "slow" LTG seems to provide an important regulatory brake on the rapid enzyme. Single-molecule fluorescent tracking of these enzymes was used to develop a model for their interaction with PG that mimics their observed activity. The rate of PG synthesis activity was also shown to impact the rate of PG degradation, suggesting potential interplay between the synthetic and degradative enzymes.

Strengths:

The genetic analysis to identify sporulation-linked LTGs and their effects on growth, sporulation, and spore properties was well done and productive. The fluorescence microscopy to track LTG mobility, presumably tied to activity, produced a convincing argument about the mechanism of regulation of one LTG by another.

Weaknesses:

While the impact of LTGs on sporulation was clearly demonstrated, the PG analysis that resulted from the study of LTGs raised some important unanswered questions. The analyses suggest that the PG is degraded to quite small fragments, which would normally be lost during the purification of PG. How these small fragments were thus detected is unclear, and this suggests a more complex story concerning PG metabolism during sporulation. An anti-PG antibody is used to quantify PG in the spores, but it is not made clear what the specificity of this antibody is, and thus whether it would recognize the LTG -altered PG of the spore. The authors suggest a "new mechanism of sporulation" when they have actually simply identified an important factor (PG degradation by LTGs) within a complex "process of sporulation".

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Details on places in the text that could be improved:

(1) Line 77: more appropriate "homologs of the sporulation genes"

Corrected.

(2) Line 137-139. Where are the 14 KO mutant data? Only showing the ones with phenotypes? Needs a citation if published elsewhere.

The phenotypes of other mutants are shown in the new Fig. S1.

(3) Line 144 and later: why "ORF"? Are they not homologous to Ltg's??. Also, the orf designation is not in the figure, so it is hard to follow the data the authors are presenting.

Following the reviewer’s recommendation, we deleted “ORF” and presented the ORF designation in the figure legend.

(4) Figure 2B - add the y-axis legend to the figure (also S1B).

Added

(5) Line 156: What does "same below" mean?

Deleted “same below”.

(6) Line 161: emtA is not indicated on Figure 2D - can you reword or add to the legend??

Changed to “MltE, an LTG encoded by Escherichia coli emtA”

(7) Line 163: "opposite roles" could be a bit better defined... do you mean ltgA induces rounding and ltgB delays rounding???

We agree with the reviewer. “opposite” was replaced by “different”.

(8) Line 172: But did the ltgA mutant make the same number of viable spores (not just FBs)? Can't conclude "the slow sporulation pathway only requires ltgB" if ltgA mutant was not tested.

We agree with the reviewer. In the revised manuscript, we quantified the starvation-induced spores and found that both LtgA and LtgB are required for forming mature spores. The data are shown in Figure 3 and supplement figures.

(9) Line 202: might be appropriate to indicate the degree of homology (% identity over protein length).

Added.

(10) Line 178, 181, and thereafter: DIC microscopy.

Corrected.

(11) Line 182/183 and 190: What is the basis for the conclusion that "polysaccharides sustained unflattened structures"??

We changed the description to “likely due to the deposition of spore coat polysaccharides that sustained unflattened cell structures (Wartel et al., 2013; Holkenbrink et al., 2014).”

(12) Line 185/186: What is the basis for the conclusion that ltgB only contained a small amount of PG? If it is based on reduced intensity, it is not obvious from the images presented. Quantitative analysis would better support this statement.

We changed the description to “Sacculi from the ∆ltgB pseudospores still contained PG but showed lower fluorescence intensity”. We also updated the figure to show the typical fluorescence intensities.

(13) Figure legend 3 line 245/6: It is not totally clear what the difference is, in that the white arrows are pointing to in ltgA vs ltgB mutant. Is it the proportion of large spherical objects in ltgB? What is the significance of the puncta in A vs. B?

We updated the description in the text as “While the remaining PG sacculi of these pseudospores were largely spherical, they lost integrity during purification, with many sacculi displaying irregular shapes in the fluorescence channel (Figure 4A).” In the figure legend, we changed the description to White arrows point to the sacculi in irregular shapes.”

(14) Line 190-192. Again, not really seeing what the authors are identifying to conclude "irregular shapes and ruptures" could authors pin-point more specifically and quantify such structures?

We updated the description in the text as “While the remaining PG sacculi of these pseudospores were largely spherical, they lost integrity during purification, with many sacculi displaying irregular shapes in the fluorescence channel (Figure 4A).”

(15) Line 194: if the authors want to make such a strong conclusion, they really should demonstrate this. Anti-spore coat antibodies are available in the field. Or take out these statements.

We took this statement out.

(16) Line 210: either they lack PG, OR the antibody can't gain access? How to conclude both? Might be more accurate to say "although we can't rule out that the polysaccharide spore coat prevents access to PG"

Following the reviewer’s recommendation, we changed the description to “These fruiting body spores lacked PG-specific fluorescence (Fig. 3A), consistent with their markedly reduced PG content (Fig. 1). However, we cannot exclude the possibility that the polysaccharide spore coats (Voelz & Dworkin, 1962) hinder antibody access to PG.

(17) Line 217-19: This is often stated in the literature, but full glycerol-induced spore maturation requires much more than 2 hours (note the authors are using O/N glycerol induction in Figure 1). And the long starvation-induced sporulation process is likely due to the differential start of sporulation, because the cells don't all enter the aggregate at the same time.... so they are not triggered to induce sporulation at the same time.

We agree with the reviewer on the first statement and changed “two hours” to “four hours”. For the second statement, we do not completely agree. Much research showed that spore development in fruiting bodies is well synchronized, for example, in (Dworkin & Voelz, 1962), starvation-induced spores did not at 48 h.

(18) Line 222: What is the evidence that it is really variation in production from the van promoter? Could it also just be due to differences in the cell cycle in the population? The authors may be correct, but should be less definitive about those conclusions since they haven't measured LtgA levels directly.

We agree with the reviewer. We changed the description to, “Upon induction with 200 μM vanillate, cells exhibited heterogeneous morphology, likely resulting from variations in LtgA expression or differences in cell cycle stages within the population.” Following this suggestion, we also changed the description on murA overexpression, “Similar to the cells that overexpressed LtgA (Fig. 3B), this heterogeneity likely reflects variable murA induction or the unsynchronized growth stages within the population.”

(19) Line 224: Where is the over-expressed LtgA data in Figure 2A? Is it that the authors are referring to over-expressed murA data, and the point is that overexpression of this kind of gene can lead to veg cell rounding? If the latter, this should be specifically stated in the text.

We apologize for this mistake. The data were shown in Fig. 3B, rather than Fig. 2A, 3B.

(20) Line 224: Did the authors test that LtgB is stably overproduced in veg cells? It could be that LtgB is turned over while LtgA is not. From the purified protein blot in Figure 3C, it does look like LtgB contains a degradation product (which is perhaps inhibiting the LtgB activity).

(21) Line 229: AgmT? Do the authors mean Ltg?

Corrected.

(22) Line 238: Is "rate" the right word? Technically, kinetics haven't been measured. Could state "LtgB less active" or "less efficient"?

Changed.

(23) Line 255: "ltgB shows a slight increase in expression during slow sporulation". Do the authors mean over the entire dev time course or specifically during the sporulation phase (which will be different timing for DZ2 vs DK1622)?

We clarified the description as “Consistent with its role in PG degradation, in a microarray-based transcriptome analysis, ltgA transcription was found to increase about twofold during rapid sporulation (4 h) but remain unchanged during slow sporulation (96 h). in the closely related DK1622 strain (Muller et al., 2010). Conversely, ltgB expression gradually rises during slow sporulation, reaching 1.8 times the vegetative level at 96 h, while remaining stable during rapid sporulation (Muller et al., 2010, Munoz-Dorado et al., 2019).”

(24) Line 278: This needs to be corrected. The authors have shown that production of fruiting bodies is not affected by the fusions. (also in Fig. S1 legend). This is really not the same as sporulation efficiency.

We changed the description to “the PAmCherry tags did not affect the formation of either glycerol-induced spores or starvation-induced fruiting bodies”.

(25) Figure S1A: panel A: Is there a deg product partially cut off at the bottom of the gel? (or is this a non-specific cross-reactive band). What is the predicted molecular mass for both proteins with the PAmCherry fusion? What conditions were these lysates generated from: veg prior to glycerol induction? I understand there are probably no antibodies available to LtgA or B, but it is important to note that it is not possible to know if there is simultaneously wt LtgA or B produced (by cleavage and degradation of the mCh fusion). Panel B: Were the differences in l/w between wt and the fusion strains tested to see if there really were no significant differences? Please state in the text (it looks like the data variance is higher in the fusion strains relative to the wt at 1 hr).

The bands at the bottom of the gel are the running front that appear in both lanes. They are not mCherry because there estimated molecular weight is much lower than that of mCherry (26.3 kDa). To avoid confusion, we cut these bands from the figure. The expression of both fusion proteins was detected from vegetative cells, which was clarified in the legend. The predicted molecular weights of them were provided in the legend too.

(26) Figure S1C: The ability to make fb is not the same as the production of spores. The authors should test the number of viable spores produced under starvation conditions, if they want to state starvation-induced sporulation is not affected by the fusions.

We agree with the reviewer. We quantified starvation-induced sporulation in the revised manuscript.

(27) Line 399: Figure S2 looks at fb formation in the absence of vanillate, not sporulation.

We changed the description to “cells grown without vanillate progressed normally through glycerol-induced sporulation and starvation-induced fruiting body formation”.

We also quantified starvation-induced sporulation in the revised manuscript.

(28) Line 281: How many fold is the ltgA transcript reduced compared to the ltgB? (i.e., If it is 1.2 fold reduced that may not be as worth mentioning as if it was 5-10 fold reduced).

ltgA transcription in vegetative cells was detected in a microarray (Muller et al., 2010) but not reported in RNAseq (Munoz-Dorado et al., 2019), significantly different from that of ltgB. We pointed this out in the revised manuscript.

(29) Figure 4B legend line 358. Define D (should it be italicized?).

Corrected.

(30) Line 363: Please define how significance was calculated. Is this the p-value?

We deleted the word “significant”.

(31) Line 298: How do the authors know immobility is from binding to PG? Provide a reference if this is well-known.

The rationale and references have been mentioned at the beginning of this section.

(32) Line 302 and thereafter: suggest "2.62 × 10-2 (plus minus) 2.0 × 10-3 μm2/s" is presented as "2.62 (plus minus) 0.20 × 10-2 μm2/s" for easier reading; switch to past tense (Were not are).

Changed following the reviewer’s recommendation.

(33) Line 310: "suggest" not "indicate", because binding of PG was directly tested.

Corrected.

(34) Line 341: "confirming it restricts access" seems very strong wording. Suggest: may compete with.

Changed.

(35) Line 392: SOME fb are larger- many are significantly smaller.

Because fruiting body sizes do not reflect sporulation efficiency, we removed this description.

(36) Figure S2 legend. Leaky expression; or on fruiting body formation.

Corrected.

(37) Line 436: stationary phase cells decrease PG synthesis- does this increase PG degradation? Perhaps it does lead to the death phase, which is striking in M. xanthus....

How do cells die, either through death phase or under antibiotic stresses, is not well understood (Baquero & Levin, 2021) Very likely, cell death is due to the accumulation of oxidative damages (Kohanski et al., 2007) and cell lysis could be a byproduct of cell death, when cells lose control of the enzymes that break PG. While cell death is a great topic to investigate, it is beyond the scope of this study.

(38) Line 507: washed.

Corrected.

(39) Line 524 (514 [sic]) and thereafter: sacculi.

Corrected.

(40) Line 587: reference for cell lysis procedure?

The procedures of cell lysis, column loading and elusion were described in details “…cells were harvested by centrifugation at 6,000 × g for 20 min and lysed by sonication in buffer A (20 mM Tris-HCl pH 8.0, 200 mM NaCl), (Nan et al., 2010, Nan et al., 2006). Proteins were loaded to an NGC™ Chromatography System (BIO-RAD) and 5-ml HisTrap™ columns (Cytiva) and eluted by buffer B (20 mM Tris-HCl pH 8.0, 200 mM NaCl, 500 mM immidazole) (Pogue et al., 2018, Nan et al., 2010).”

(41) Line 269: by microscopy (or "under THE microscope").

Corrected.

(42) Line 271: THE cell/PG.

“The” added.

(43) Line 603: OR not and.

Corrected.

(44) Line 497 (and elsewhere): Is it really CFU? If determined by OD, then not technically CFU because some cells will not grow into colonies.

We agree with the reviewer. Cell concentrations were determined by OD. “CFU” was deleted.

(45) Line 506: washed.

Same as recommendation (38). Corrected.

(46) Line 521: min.

Corrected.

(47) Line 532: How were peaks assigned?

We described peak assignment in details in the revised manuscript, “Muropeptides were assigned based on: (i) accurate mass matching to theoretical monoisotopic masses of expected M. xanthus PG building blocks (Bui et al., 2009, White et al., 1968) and (ii) comparison of retention times with those reported in previous analysis with similar PG compositions.”

Reviewer #2 (Recommendations for the authors):

(1) If almost all the muropeptides detected in spores are anhydro products of LTGs, then it might be expected that these are all very small peptidoglycan fragments in the spores. If the anhydro units were at the ends of short PG chains, then muramidase digestion would release similar amounts of non-anhydro products, but none are detected. So, is muramidase digestion doing anything to the PG derived from the spores? Is muramidase digestion required to observe the spore muropeptide pattern? A control sample in which muramidase digestion is omitted would answer these questions.

Muramidase digestion is essential for solubilizing PG into different muropeptides for UPLC analysis. In each sample, both anhydro and non-anhydro products were detected. In our original submission, we pointed out that “The two spore types showed similar profiles of a discernible presence of muropeptides that resembled those found in vegetative cells, albeit in significantly reduced quantities (Fig. 1).” We clarified the PG analysis. “The purification procedure yields only sedimentable PG, as all soluble fragments are removed during the washing steps. The resulting sacculi were then digested with muramidase, and the solubilized muropeptides were analyzed by UPLC (see Materials and Methods). The chromatograms in Fig. 1 reflect the muropeptides released specifically from the sedimented sacculus fraction.” Because muramidase release polysaccharides or disaccharides, so it’s digestion does not release nonanhydro GlcNAc species, which is why we did not see equal amounts of anhydro and non-anhydro products. In our case, over 90% of the polysaccharides and disaccharides contain Anhydro-MurNAc. The dominance of anhydro products indicates that LTGs cut very frequently on glycan chains.

(2) This also raises the question of how these very small peptidoglycan fragments are even retained in the spores. They would be expected to be lost during spore purification or during PG purification prior to muramidase digestion. How do you even purify sacculi when the spores have no PG chains? One could theorize that the polysaccharide coats hold everything in, but then the PG would be protected from muramidase digestion.

We thank the reviewer for this question. Our analysis suggests that M. xanthus spores do not lack PG entirely but retain a residual PG mesh that is still crosslinked. Such crosslinked material sediments during PG purification and remains accessible to muramidase, which hydrolyses internal glycosidic bonds and releases anhydro muropeptides. Non-crosslinked fragments would indeed be washed away during purification steps, so the detected muropeptides reflect the structure of this residual sacculus (Fig. 1).

(3) What is the anti-PG antibody recognizing, the glycan backbone, the peptide side chain, or both? Can the antibody recognize the very short anhydro-containing disaccharides proposed to be predominant in spores? If not, then the PG quantification using the antibody is not accurate.

The structures recognized by the anti-PG antibodies are unknown. Our samples do not contain small degradation products, which was clarified in the revised manuscript, “To answer this question, we purified cell sacculi and used immunofluorescence and an anti-PG serum (de Pedro et al., 1997) to visualize the remaining PG.” We used immunofluorescence to display the PG scaffolds remained in each sample and we did not perform any quantitative analysis based on the images.

(4) Lines 325-327: This is a speculative conclusion and should be stated as such, i.e., "may control the pace.." This conclusion could be somewhat more strongly stated at the end of the next section, around lines 345-350.

Moved following the reviewer’s recommendation.

(5) Lines 403-404: This first sentence of the discussion seems completely dissociated from the topic of the paper; it should be deleted.

Deleted.

(6) Lines 404-405. I am not convinced that these findings "elucidate a new mechanism of sporulation." The study was undertaken because PG degradation was already tied to sporulation in a previous study. Furthermore, I am not sure that PG degradation is a "mechanism of sporulation." It is clearly an important step in sporulation of this species, but is it the driving "mechanism"?

We tuned down our statement as “Our findings demonstrate that M. xanthus, a nonfirmicute bacterium, relies on PG degradation to change cell shape during sporulation”.

(7) Lines 510-516 describe PG purification from vegetative cells. How was this process modified for spores?

We apologize for the confusion. The process was clarified as “For PG analysis, samples were processed as previously described for Gram-negative bacteria (Alvarez et al., 2016; Desmarais et al., 2013). Vegetative cells were harvested at mid-stationary phase by centrifugation (30 min, 8,000 g). Vegetative cells and purified spores (as described in the previous section) were resuspended…”.

Alvarez, L., Hernandez, S.B., de Pedro, M.A., and Cava, F. (2016) Ultra-Sensitive, High-Resolution Liquid Chromatography Methods for the High-Throughput Quantitative Analysis of Bacterial Cell Wall Chemistry and Structure. Methods Mol Biol 1440: 1127.

Baquero, F., and Levin, B.R. (2021) Proximate and ultimate causes of the bactericidal action of antibiotics. Nat Rev Microbiol 19: 123-132.

Bui, N.K., Gray, J., Schwarz, H., Schumann, P., Blanot, D., and Vollmer, W. (2009) The peptidoglycan sacculus of Myxococcus xanthus has unusual structural features and is degraded during glycerol-induced myxospore development. J Bacteriol 191: 494505.

de Pedro, M.A., Quintela, J.C., Holtje, J.V., and Schwarz, H. (1997) Murein segregation in Escherichia coli. J Bacteriol 179: 2823-2834.

Desmarais, S.M., De Pedro, M.A., Cava, F., and Huang, K.C. (2013) Peptidoglycan at its peaks: how chromatographic analyses can reveal bacterial cell wall structure and assembly. Mol Microbiol 89: 1-13.

Dworkin, M., and Voelz, H. (1962) The formation and germination of microcysts in Myxococcus xanthus. J Gen Microbiol 28: 81-85.

Holkenbrink, C., Hoiczyk, E., Kahnt, J., and Higgs, P.I. (2014) Synthesis and assembly of a novel glycan layer in Myxococcus xanthus spores. J Biol Chem 289: 32364-32378.

Kohanski, M.A., Dwyer, D.J., Hayete, B., Lawrence, C.A., and Collins, J.J. (2007) A common mechanism of cellular death induced by bactericidal antibiotics. Cell 130: 797-810.

Lee, T.K., Meng, K., Shi, H., and Huang, K.C. (2016) Single-molecule imaging reveals modulation of cell wall synthesis dynamics in live bacterial cells. Nature communications 7: 13170.

Muller, F.D., Treuner-Lange, A., Heider, J., Huntley, S.M., and Higgs, P.I. (2010) Global transcriptome analysis of spore formation in Myxococcus xanthus reveals a locus necessary for cell diberentiation. BMC Genomics 11: 264.

Munoz-Dorado, J., Moraleda-Munoz, A., Marcos-Torres, F.J., Contreras-Moreno, F.J., MartinCuadrado, A.B., Schrader, J.M., Higgs, P.I., and Perez, J. (2019) Transcriptome dynamics of the Myxococcus xanthus multicellular developmental program. Elife 8.

Nan, B., Liu, X., Zhou, Y., Liu, J., Zhang, L., Wen, J., Zhang, X., Su, X.D., and Wang, Y.P. (2010) From signal perception to signal transduction: ligand-induced dimeric switch of DctB sensory domain in solution. Mol Microbiol 75: 1484-1494.

Nan, B., Zhou, Y., Liang, Y.H., Wen, J., Ma, Q., Zhang, S., Wang, Y., and Su, X.D. (2006) Purification and preliminary X-ray crystallographic analysis of the ligand-binding domain of Sinorhizobium meliloti DctB. Biochim Biophys Acta 1764: 839-841.

Pogue, C.B., Zhou, T., and Nan, B. (2018) PlpA, a PilZ-like protein, regulates directed motility of the bacterium Myxococcus xanthus. Mol Microbiol 107: 214-228.

Ramirez Carbo, C.A., Faromiki, O.G., and Nan, B. (2024) A lytic transglycosylase connects bacterial focal adhesion complexes to the peptidoglycan cell wall. Elife 13.

Ramírez Carbó, C.A., and Nan, B. (2026) Using Single-Particle Fluorescence Microscopy to Quantify Substrate Binding of Peptidoglycan-Modification Enzymes. Bio-protocol 16: e5696.

Voelz, H., and Dworkin, M. (1962) Fine structure of Myxococcus xanthus during morphogenesis. J Bacteriol 84: 943-952.

Wartel, M., Ducret, A., Thutupalli, S., Czerwinski, F., Le Gall, A.V., Mauriello, E.M., Bergam, P., Brun, Y.V., Shaevitz, J., and Mignot, T. (2013) A versatile class of cell surface directional motors gives rise to gliding motility and sporulation in Myxococcus xanthus. PLoS Biol 11: e1001728.

White, D., Dworkin, M., and Tipper, D.J. (1968) Peptidoglycan of Myxococcus xanthus: structure and relation to morphogenesis. J Bacteriol 95: 2186-2197.

Yang, X., McQuillen, R., Lyu, Z., Phillips-Mason, P., De La Cruz, A., McCausland, J.W., Liang, H., DeMeester, K.E., Santiago, C.C., Grimes, C.L., de Boer, P., and Xiao, J. (2021) A two-track model for the spatiotemporal coordination of bacterial septal cell wall synthesis revealed by single-molecule imaging of FtsW. Nat Microbiol 6: 584-593.

Zhang, H., Venkatesan, S., Ng, E., and Nan, B. (2023) Coordinated peptidoglycan synthases and hydrolases stabilize the bacterial cell wall. Nature communications 14: 5357.

Version 2

Abstract

Many bacteria form spores to endure unfavorable conditions. While Firmicutes generate endospores through cell division, sporulation in non-Firmicutes remains less understood. The Gram-negative bacterium Myxococcus xanthus undergoes sporulation through two distinct mechanisms: rapid sporulation triggered by chemical induction and slow sporulation driven by starvation, both occurring independently of cell division. Instead, these processes depend on the complete degradation of the peptidoglycan (PG) cell wall by two lytic transglycosylases (LTGs), LtgA and LtgB. Remarkably, LtgB programs the pace of PG degradation by LtgA during rapid sporulation, ensuring a controlled process that prevents abrupt PG breakdown and the formation of non-resistant pseudospores. In addition to regulation between LTGs, PG degradation is also influenced by its synthesis; cells exhibiting increased muropeptide production often circumvent sporulation. These findings not only reveal novel mechanisms of bacterial sporulation but also shed light on the regulatory network governing PG dynamics.

Introduction

Spores are metabolically dormant cells that can survive unfavorable conditions, including extremes of temperature, desiccation, and ionizing radiation (Huang & Hull, 2017, Hutchison et al., 2014). Sporulation, the process of spore development, is a strategy utilized by a wide variety of organisms, from bacteria and protozoa to fungi and plants. In addition to protective spore coats, spore resilience relies on cytosol dehydration and DNA compaction—key processes that decrease cell volume and alter cell shape (Higgins & Dworkin, 2012). Thus, morphological differentiation is a hallmark of sporulation. In bacterial spore formers, as peptidoglycan (PG) cell walls largely determine cell morphology, their sporulation requires profound PG remodeling.

PG is a rigid, mesh-like macromolecule that is composed of glycan strands of repeating units of N-acetyl glucosamine (GluNAc)-N-acetyl muramic acid (MurNAc) crosslinked by peptides (Egan et al., 2020). PG encloses the entire cytoplasmic membrane, and its rigidity provides mechanical support against osmotic stress. For this reason, PG is an essential structure for most bacteria and a major target for antibacterial treatments. Sporulation provides invaluable opportunities for understanding the dynamics of PG, which is controlled by multiple synthases and hydrolases. PG synthases include glycosyltransferases (GTases) that polymerize glycan strands and transpeptidases (TPases) that form peptide crosslinks (Egan et al., 2020). PG hydrolases, also known as autolysins, include glycosidases that break the glycan strands, and amidases and endo/carboxypeptidases that cleave the peptide crosslinks (Egan et al., 2020, Rohs & Bernhardt, 2021, van Heijenoort, 2011).

Firmicutes, including Gram-positive bacteria like Bacilli and Clostridia, produce endospores. Their morphological transition from rod-shaped vegetative cells to ovoid spores occurs through asymmetric cell division, resulting in the formation of a smaller forespore and a larger mother cell. Eventually, the forespore becomes an oval endospore after being engulfed by the mother cell (Higgins & Dworkin, 2012, McKenney & Eichenberger, 2012). A mature endospore contains two PG layers: the germ cell wall, derived from the forespore’s original PG, and a thickened PG cortex deposited by the mother cell (Popham & Bernhards, 2015). While a few Gram-negative spore formers also belong to the Firmicutes phylum, they share conserved sporulation genes with Bacilli and Clostridia, which suggests similar sporulation mechanisms (Tocheva et al., 2011, Yutin & Galperin, 2013).

Sporulation by non-firmicutes bacteria has been largely overlooked. Myxobacteria, a group of Gram-negative, non-firmicutes spore formers, were first assigned to the phylum δ-proteobacteria, but recently reclassified into the newly established phylum Myxococcota (Waite et al., 2020). Myxococcus xanthus is a model organism of myxobacteria. Rod-shaped M. xanthus cells can undergo sporulation via two distinct pathways, both resulting in spherical spores (Kroos et al., 2025). First, in response to certain chemical signals, such as glycerol and dimethyl sulfoxide (DMSO), individual M. xanthus cells can rapidly transition into isolated spherical spores in aqueous environments within hours (Dworkin & Gibson, 1964). Second, millions of cells can aggregate on solid starvation media and develop into spore-filled fruiting bodies, a process that takes a few days to complete (O’Connor & Zusman, 1988). Both sporulation pathways are programmed, tightly controlled processes that involve over 1,000 genes (Muller et al., 2010, Munoz-Dorado et al., 2019). In contrast to endospore formation, cell division is not involved in either the M. xanthus sporulation mechanisms, and M. xanthus lacks the homologs of the sporulation genes in Firmicutes (Aramayo & Nan, 2022). Thus, sporulating M. xanthus must accomplish the rod-to-sphere transition through yet to be discovered PG remodeling mechanisms, which provide invaluable opportunities to investigate PG dynamics (Zhang et al., 2021).

Distinct from Bacilli and Clostridia, M. xanthus spores lack cortex PG (Bui et al., 2009, Zhang et al., 2020, Voelz & Dworkin, 1962). Rather, their resistance is derived from polysaccharide coats (Kottel et al., 1975, Perez-Burgos et al., 2020, Saidi et al., 2021, Muller et al., 2012). Sporulating cells that fail to deposit coat polysaccharides on their surfaces produce spherical pseudospores that lack resistance (Wartel et al., 2013, Zhang et al., 2020, Holkenbrink et al., 2014). In a pioneering study, Bui et al. did not detect muropeptides in glycerol-induced spores, indicating that M. xanthus degrades its PG during rapid sporulation (Bui et al., 2009). However, it remains unclear whether starvation-induced spores within fruiting bodies still retain PG and how M. xanthus orchestrates programmed morphological changes.

In this report, we used ultra-performance liquid chromatography (UPLC) to demonstrate that M. xanthus degrades PG in both sporulation pathways. Through mutagenesis studies, we identified LtgA and LtgB, two lytic transglycosylases (LTGs) essential for these sporulation pathways. Remarkably, LtgB regulates the pace of PG degradation by LtgA during rapid sporulation, preventing the formation of non- resistant pseudospores due to abrupt PG breakdown. This research not only uncovers novel mechanisms of sporulation in non-firmicutes but also highlights the crucial role of cross-regulation between PG hydrolases in maintaining cell integrity.

Results

PG is degraded in both spore types

To investigate if starvation-induced spores in M. xanthus fruiting bodies retain PG, we broke the fruiting bodies after 120 h of starvation and purified saculli from spores. For comparison, we also purified the sacculi from vegetative cells and glycerol-induced spores. The purification procedure yields only sedimentable PG, as all soluble fragments are removed during the washing steps. The resulting sacculi were then digested with muramidase, and the solubilized muropeptides were analyzed by UPLC (see Materials and Methods). The chromatograms in Figure 1 reflect the muropeptides released specifically from the sedimented sacculus fraction. The two spore types showed similar profiles of a discernible presence of muropeptides that resembled those found in vegetative cells, albeit in significantly reduced quantities (Figure 1). Importantly, the most abundant muropeptides identified in the vegetative cells, MurNAc-tetrapeptide monomer (M4) and MurNAc-tetrapeptide dimer (D44), were nearly absent in both spore types (Figure 1). Such residual muropeptides in both spore types are inadequate for forming continuous PG layers (Bui et al., 2009). This observation indicates that spore development in both pathways involves the breakdown of the vegetative PG cell wall.

Mature M. xanthus spores induced by either glycerol or starvation do not contain significant muropeptides.

UPLC muropeptides profiles of vegetative cells, glycerol- and starvation-induced spores indicate that the major muropeptides species, M4 and D44, in vegetative cells are diminished in both spore types, while spores are enriched in anhydro- muropeptides (Anh). Only muropeptides that could be confidently identified by MS/MS analysis are labelled. The characteristic peaks are labeled as follows: M, monomeric muropeptide (uncrosslinked); D, dimeric muropeptide (crosslink connecting two muropeptides), T, trimeric muropeptide (crosslink connecting three muropeptides). Numbers refer to the status of the peptide side chain (3, tripeptide; 4, tetrapeptide). Red characters mark the muropeptides only detected in spores.

Despite the overall decline in muropeptides, anhydro-muropeptides markedly increased in both spore types, comprising over 90% of the total muropeptides (Figure 1). Especially, anhydro-MurNAc-tetrapeptide trimer (T444Anh) that was only detected in trace amounts in vegetative cells, became a prominent muropeptide in both spore types (Figure 1). Moreover, new anhydro-muropeptides, including anhydro, anhydro- MurNAc-tetrapeptide dimer (D44Anh,Anh), anhydro, anhydro-MurNAc-tetrapeptide trimer (T444Anh,Anh), and an anhydrodimer without a NAG (D44Anh(-NAG)), were only present in spores (Figure 1). Because anhydro-muropeptides are the signature products of LTGs (Dik et al., 2017, Williams et al., 2018), their abundance in spores indicates that certain LTGs must play essential roles in M. xanthus sporulation.

M. xanthus sporulation requires two LTGs

The genome of M. xanthus encodes 14 putative LTGs (Aramayo & Nan, 2022, Ramirez Carbo et al., 2024). We imaged the cells of the 14 knockout mutants (Ramirez Carbo et al., 2024) after 6 h of glycerol induction and identified two mutants that displayed abnormality in sporulation (Figure 2, Figure 2 – figure supplement 1). We then imaged the wild-type and mutant cells at different time points after glycerol induction and used their length/width (L/W) ratios to monitor potential defects in the sporulation process. Wild-type cells initiated sporulation within 1 h of glycerol induction, which was evidenced by the increase of cell width and decrease of cell length (Figure 2A). Consistent with the change in cell morphology, their L/W ratios decreased continuously and stabilized after 2 h of induction, when sporulation completed. In contrast, the cells that carry the deletion of K1515_20820 (MXAN_RS16290) (Aramayo & Nan, 2022) were able to shorten cell length slightly but retained rod shape after prolonged induction (Figure 2B, Figure 2 – source data 1).

LtgA and LtgB are required for glycerol-induced sporulation, whereas MurA overexpression inhibit sporulation.

A) Bright field images of cells at different time points after glycerol-induction. Black arrows point to lysing cells. B) Quantitative analysis of glycerol- induced sporulation using the length/width ratio (L/W) of cells. Whiskers indicate the 25th - 75th percentiles and red dots the median. The total number of cells analyzed is shown on top of each plot. C) LtgA and LtgB are homologous to E. coli MltE. The asterisk marks the conserved active site. Other putative LTGs do not affect either sporulation pathway significantly, which is shown in Figure 2 – figure supplement 1.

Surprisingly, another mutant that carries the deletion of K1515_17460 (MXAN_RS19615) started to lose rod shape immediately after glycerol induction and became spherical within 1 h (Figure 2A, 2B, Figure 2 – source data 1). However, different from the wild-type spores that appeared dark and heterogeneous under differential interference contrast (DIC) microscopy, the spheres formed by this mutant appeared bright and homogenous, similar to the pseudospores from the ΔaglQS mutant that lack the motor for depositing spore coat polysaccharides onto cell surfaces (Zhang et al., 2020) (Figure 2A). To test whether these spheres are real spores, we subjected them to sonication and quantified their survival rate using a Helber bacterial counting chamber. After sonication, while 91 ± 6% (calculated from three independent experiments, n > 1,000) of the wild-type spores appeared intact, only 3 ± 1% of the spheres produced by the K1515_17460 deletion mutant remained. Thus, this mutant indeed formed pseudospores that lacked the resistance against sonication. As the products of K1515_20820 and K1515_17460 show homology to MltE, an LTG encoded by Escherichia coli emtA (49.0% and 38.4% sequence identity, respectively) (Figure 2C), we named them ltgA and ltgB, respectively. Both LtgA and LtgB are required for forming resistant spores via the rapid sporulation pathway, albeit playing different roles in the rod-to-sphere transition.

To determine if these LTGs are also required for starvation-induced sporulation, we grew them in rich liquid media and spotted cells on solid starvation (CF) media. After 96 h of incubation, both mutants produced dark fruiting bodies on the agar surface that were comparable to those of the wild-type (Figure 3A). To test if these mutants form starvation-induced spores, we incubated cells in static liquid CF medium using polystyrene Petri dishes for 96 h, scraped the cells attached to the dishes, suspended them in 1 ml water, subjected them to sonication, then plated them on solid rich media. After 96 h of incubation, while the wild-type fruiting bodies produced colonies of 3389 ± 1141 cfu/ml (n = 3), the ΔltgA and ΔltgB cells failed to form colonies after five days of incubation (Figure 3B, Figure 3 – source data 1), indicating that both LtgA and LtgB are essential for forming starvation-induced spores.

LtgA, LtgB, and MurA regulate starvation-induced sporulation.

A) The absence of LtgA and LtgB, and the overexpression of MurA do not affect the formation of fruiting body– like aggregates on starvation agar. B) Both LtgA and LtgB are essential for forming starvation- induced spores, while MurA overexpression negatively regulates this process. Sonication- resistant spores were counted as colony formation units (cfu) after 96 h of incubation on CYE agar. Data are shown as mean ± standard deviation (n = 3).

LtgA and LtgB play distinct roles in different sporulation pathways

Do LtgA and LtgB degrade PG during sporulation? To answer this question, we purified cell sacculi and used immunofluorescence and an anti-PG serum (de Pedro et al., 1997) to visualize the remaining PG. The sacculi of vegetative cells from the wild- type, ΔltgA, and ΔltgB strains were not visible under DIC microscopy, likely due to their flattened structures minimizing light interference. However, PG from all three strains was readily detected in the fluorescence channel (Figure 4A). After 6 h of glycerol-induction, the sacculi of both the wild-type and ΔltgA cells remained visible under DIC microscopy (Figure 4A), likely due to the deposition of spore coat polysaccharides that sustained unflattened cell structures (Wartel et al., 2013, Holkenbrink et al., 2014). After the sacculus purification process, only background PG signals were detected in glycerol-induced wild-type spores (Figure 4A). In contrast, the ΔltgA cells retained PG in substantial quantities (Figure 4A). Sacculi from the ΔltgB pseudospores still contained PG but showed lower fluorescence intensity (Figure 4A). While the remaining PG sacculi of these pseudospores were largely spherical, they lost integrity during purification, with many sacculi displaying irregular shapes in the fluorescence channel (Figure 4A). Thus, compared to LtgA, the enzymatic activity of LtgB plays a more limited role in PG degradation during chemical-induced sporulation. Collectively, LtgB appears to be a pace-keeper that prevents abrupt PG degradation. Different from the induced wild-type and ΔltgA cells, sacculi from the ΔltgB pseudospores were undetectable under DIC microscopy (Figure 4A), reflecting their flattened shapes. This phenotype suggests that further acceleration of the morphological transitions associated with rapid sporulation may impede the formation of a rigid polysaccharide coat.

LtgA and LtgB play distinct roles in the two sporulation pathways.

A) While LtgA is required for PG degradation during glycerol-induced sporulation, LtgB is the major LTG for forming starvation-induced spores. PG was detected using an anti-PG serum and a fluorescence-conjugated secondary antibody. PG sacculi were purified from cells after 6 h and 120 h of glycerol-induced and starvation-induced sporulation, respectively. Flattened sacculi are not visible under bright field microscopy. White arrows point to the sacculi in irregular shapes. BF, bright field. B) The overexpression of LtgA, but not LtgB, collapses rod-shape in vegetative cells. C) Purified LtgA and LtgB solubilize dye-labeled PG sacculi at different rates. Lysozyme and buffer serve as the positive and negative controls, respectively. Absorption at 595 nm was measured after 18 h incubation at 25 ⁰C. Data are presented as mean values ± SD from three technical replicates. The inset shows purified LtgA and LtgB in a Coomassie stained gel. Scale bars, 5 μm.

To determine the roles of LtgA and LtgB in slow sporulation, we first scraped the wild-type fruiting bodies from CF agar surface (Figure 2C), dispersed spores by sonication, purified their sacculi, and visualized PG using immunofluorescence. Spherical spores from the wild-type cells were visible under DIC microscope, indicating that they retained unflattened shapes. These fruiting body spores lacked PG-specific fluorescence (Figure 4A), consistent with their markedly reduced PG content (Figure 1). However, we cannot exclude the possibility that the polysaccharide spore coats (Voelz & Dworkin, 1962) hinder antibody access to PG. Second, we investigated if ΔltgA and ΔltgB cells degraded PG during starvation-induced sporulation. Because the aggregates formed by ΔltgA and ΔltgB cells on CF agar did not contain mature spores, we scrapped cell aggregates from agar surface and purified their sacculi without sonication. The ΔltgA cells formed spherical, spore-like cells void of PG (Figure 4A), indicating that while LtgB is sufficient for PG degradation during slow sporulation, LtgA is only required for spore maturation. In contrast, most of the ΔltgB cells were still rod-shaped, indistinguishable from vegetative ones (Figure 4A). These observations indicate that LtgB is an essential LTG for PG degradation during slow sporulation.

The two sporulation pathways vary greatly in duration: glycerol-induced rapid sporulation is completed within four hours, while starvation-induced slow sporulation unfolds over several days. Based on this distinction, we hypothesized that LtgA and LtgB could degrade PG at different rates. We used a vanillate-inducible promoter (Iniesta et al., 2012) to overexpress LtgA and LtgB as a merodiploids. Upon induction with 200 µM vanillate, cells overproducing LtgA exhibited heterogeneous morphology, likely resulting from variations in LtgA expression or differences in cell cycle stages within the population. Many cells lost rod shape even in the absence of glycerol, indicating that these cells over-degraded their PG by LtgA (Figure 4B). In contrast, the overexpression of LtgB using the same method did not affect the morphology of vegetative cells (Figure 4B). To test if LtgB overexpression was induced by vanillate, we added a photo-activatable mCherry (PAmCherry) tag to the overexpression constructs of LtgA and LtgB. As shown in Figure 4 – figure supplement 1, in the presence of vanillate, the expression of both LTGs increased significantly. These findings suggest that LtgB exhibits very low LTG activity in vegetative cells, as overexpression of the protein fails to induce substantial PG degradation.

To further investigate the activities of these LTGs, we expressed the periplasmic domains of wild-type LtgA (amino acids 21 – 243) and LtgB (amino acids 26 – 710) in E. coli (Figure 4C, Figure 4 – source data 1). We purified PG from wild-type M. xanthus cells, labeled it with Remazol brilliant blue (RBB), and tested if the purified LTGs hydrolyze labeled PG in vitro and release the dye (Jorgenson et al., 2014, Ramirez Carbo et al., 2024, Uehara et al., 2010). Wild-type LtgA solubilized dye-labeled PG, which absorbed light at 595 nm, demonstrating stronger hydrolytic activity than lysozyme, an enzyme that specifically cleaves β-1,4-glycosidic bonds in PG. In contrast, PG incubated with LtgB only showed minimum release of the dye, indicating slow PG hydrolysis (Figure 4C, Figure 4 – source data 1). While we cannot rule out the possibility that our purification and reaction conditions were suboptimal for LtgB, both the in vitro RBB assay and the in vivo phenotype resulting from LtgB overexpression support our hypothesis that during vegetative growth, LtgB is less efficient in PG degradation than LtgA.

LtgB regulates LtgA during glycerol-induced sporulation

Consistent with its role in PG degradation, in a microarray-based transcriptome analysis, ltgA transcription was found to increase about twofold during rapid sporulation (4 h) but remain unchanged during slow sporulation (96 h) in the closely related DK1622 strain (Muller et al., 2010). Conversely, ltgB expression gradually rises during slow sporulation, reaching 1.8 times the vegetative level at 96 h, while remaining stable during rapid sporulation (Muller et al., 2010, Munoz-Dorado et al., 2019). However, transcriptomic data alone do not account for the opposing roles of LtgA and LtgB in the rod-to-sphere transition during rapid sporulation. The resemblance in morphology between glycerol-induced ΔltgB cells and uninduced cells overproducing LtgA (Figure 4B) suggests that the slower-acting LtgB may offset the rapid activity of LtgA during sporulation, allowing sufficient time for spore maturation.

To test if LtgB regulates LtgA, we constructed a ΔltgA ΔltgB double deletion strain. Cells from this strain shortened their length slightly but failed to abolish rod shape after prolonged glycerol induction, phenocopying the sporulation defect of the ΔltgA strain (Figure 2A, 2B, Figure 2 – source data 1). Hence, LtgB is a regulator upstream of LtgA.

LtgB exhibits a response to glycerol induction earlier than LtgA

Biochemical reactions on PG are unique for the stark size difference between the enzymes and their substrates. While the enzymes are in nanometer scales, their substrates, the PG sacculi, span several micrometers. Under the microscope, PG remains stationary but PG-related enzymes are free to move. Even for M. xanthus that moves on surfaces, PG-related enzymes move at least two orders of magnitude faster than the cell/PG (Zhang et al., 2023b, Ramirez Carbo et al., 2024). Thus, when diffusive enzymes bind to PG, their mobility decreases (Lee et al., 2016, Zhang et al., 2023b). For instance, DacB, another PG hydrolase, reduces its single-particle mobility in the conditions where its activity is activated (Zhang et al., 2023b). By tracking single fluorescently-labeled enzyme particles, we can approximate their PG-binding in different physiological conditions and genetic backgrounds (Ramirez Carbo et al., 2024, Zhang et al., 2023b, Ramírez Carbó & Nan, 2026). We individually expressed PAmCherry-labeled LtgA and LtgB using their native loci and promoters. Both labeled LTGs accumulated as full- length proteins (Figure 5 – figure supplement 1A) and the PAmCherry tags did not negatively affect either glycerol or starvation-induced sporulation pathways, indicating that these fusion proteins were fully functional (Figure 5 – figure supplement 1B, C). Notably, vegetative cells produce significantly less LtgA than LtgB, consistent with the RNAseq data in which ltgA transcription was below the detection limit (Munoz-Dorado et al., 2019) (Figure 5 – figure supplement 1A).

We used a 405-nm excitation laser (0.3 kW/cm2, 0.1 s) to activate the fluorescence of a few labeled LTG particles randomly in each cell and quantified their localization using a 561-nm laser at 10 Hz using single particle tracking photo-activated localization microscopy (sptPALM, see Materials and Methods) (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013, Ramírez Carbó & Nan, 2026). As free PAmCherry particles diffuse extremely fast in the cytoplasm, entering and exiting the focal plane frequently, they usually appear as blurry objects that cannot be followed at 10 Hz close to the cell surface (Fu et al., 2018, Zhang et al., 2023b). For this reason, the noise from free PAmCherry due to potential protein degradation was negligible.

Single-particles of PAmCherry-labeled LtgA and LtgB can be categorized into two populations, immobile and mobile. The immobile particles remained within a single pixel (160 nm × 160 nm) before photobleach, and the mobile ones displayed typical diffusion (Figure 5A, Figure 5 – source data 1). For the particles that switched between mobile and immobile states, our algorithm categorized them as mobile and calculated their diffusion coefficients (D) from their entire trajectories that contained both mobile and immobile segments. Hence, binding to PG not only increases the immobile population of the enzyme particles but also decreases the D of the mobile particles. In vegetative cells where large-scale PG degradation does not occur, 23.9% (n = 1175) and 18.2% (n = 824) of LtgA and LtgB particles were immobile, respectively (Figure 5B, Figure 5 – source data 1). D values of the mobile population were (2.62 ± 0.20) × 10-2 µm2/s (n = 894) for LtgA-PAmCherry and (2.80 ± 0.25) × 10-2 µm2/s (n = 674) for LtgB- PAmCherry (Figure 5C, Figure 5 – source data 1).

LtgB regulates the PG-binding of LtgA during glycerol-induced sporulation.

A) LtgB exhibits a response to glycerol induction earlier than LtgA. Representative trajectories of LtgA and LtgB before (uninduced) and after (1 min and 30 min) glycerol induction. The overall distribution of both LTGs is displayed using the composite of 100 consecutive frames taken at 100-ms intervals. Single-particle trajectories of PAmCherry were generated from the same frames. Individual trajectories are distinguished by colors. Scale bars, 5 μm. B and C) the absence of LtgB reduces the diffusion of LtgA, which is reflected in the increase of immobile population (B) and the decrease in diffusion coefficient (D) (C), and these effects are especially prominent in the cells before (uninduced) and immediately after (1 min) glycerol induction. For each protein and condition, particles were identified from at least 100 cells and three independent experiments. The total number of particles analyzed is shown on top of each plot. Error bars were the standard derivation of 1,000 bootstrap samples and * indicates a difference of > 0.005.

We then determined how the dynamics of both LTGs respond to glycerol induction in the rapid sporulation pathway. Immediately (1 min) after adding glycerol, the dynamics of LtgA remained little changed, when 21.0% (n = 928) of particles were immobile and the D of the mobile ones was (2.53 ± 0.19) × 10-2 µm2/s (n = 733) (Figure 5B, C, Figure 5 – source data 1). In contrast, the mobility of LtgB decreased significantly, with the immobile population increased to 28.0% (n = 1359) and D of the mobile population decreased to (2.17 ± 0.28) × 10-2 µm2/s (n = 978) (Figure 5B, C, Figure 5 – source data 1). Therefore, these results suggest that upon glycerol induction, LtgB rapidly enhances its binding to PG.

We then chose 30 min after glycerol induction as a time point for rapid PG degradation, which is reflected by the dramatic decrease of L/W during the first hour of sporulation (Figure 2A, B, Figure 2 – source data 1). Compared to its inert response 1 min after glycerol induction, the mobility of LtgA particles decreased significantly at 30 min, when the immobile population increased to 31.5% (n = 1163) and D of the mobile population decreased to (1.80 ± 0.21) × 10-2 µm2/s (n = 797) (Figure 5B, C, Figure 5 – source data 1). The reduced mobility of LtgA aligns with its function as the primary LTG in glycerol-induced sporulation. Strikingly different from LtgA, as sporulation advanced to the 30-min mark, LtgB’s mobility returned to its pre-induction level, with the immobile population decreased to 10.2% (n = 1549) and D of the mobile population increased to (2.87 ± 0.28) × 10-2 µm2/s (n = 1391) (Figure 5B, C, Figure 5 – source data 1). Therefore, the PG-binding of LtgB shows a negative correlation with PG degradation. Taken together, LtgA and LtgB display opposite responses to glycerol induction, in which LtgB rapidly binds to PG before yielding to LtgA, whose PG-binding is concurrent with the rod-to-sphere transition.

LtgB blocks LtgA from binding PG in the early stage of glycerol-induced sporulation

Does LtgB’s early PG-binding suppress PG degradation by LtgA? To answer this question, we expressed LtgA-PAmCherry using the native ltgA locus and promoter in the ΔltgB background. In vegetative cells, the absence of LtgB significantly reduced the mobility of single LtgA-PAmCherry particles, suggesting that LtgB does affect LtgA’s binding to the PG (Figure 5B, C, Figure 5 – source data 1). Similarly, we expressed LtgB- PAmCherry using the native ltgB locus and promoter in the ΔltgA background. In contrast, the mobility of single LtgB-PAmCherry particles only decreased slightly in the absence of LtgA (Figure 5B, C, Figure 5 – source data 1). These results suggest that while LtgB significantly reduces LtgA’s binding to PG, probably due to its higher expression level, LtgA has only a modest impact on LtgB’s ability to bind PG.

Strikingly different from its slow response to glycerol in wild-type cells, LtgA increased its PG-binding immediately (1 min) after glycerol induction in the ΔltgB background, confirming that LtgB may compete with LtgA for access to PG in the early stage of sporulation (Figure 5B, C, Figure 5 – source data 1). In contrast, at 30 min of induction, a time point when LtgB dissociated from PG, its absence no longer affected LtgA’s PG-binding (Figure 5B, C, Figure 5 – source data 1). Similar to the observation in wild-type cells, the lack of LtgA did not affect the PG-binding of LtgB at either 1 min or 30 min after glycerol induction (Figure 5B, C, Figure 5 – source data 1). In summary, the sequential PG-binding by these two LTGs controls the pace of PG-degradation during glycerol-induced sporulation. LtgB, a slow LTG, regulates the progression of rapid sporulation by preventing LtgA, a fast LTG, from excessively degrading PG, thereby ensuring spore maturation.

Upregulated PG synthesis negatively affects PG degradation

Since LtgA is less mobile in vegetative ΔltgB cells, i.e. strongly binds to PG (Figure 5B, C, Figure 5 – source data 1), why do these cells retain their rod shape rather than losing it spontaneously before glycerol induction? In addition to PG-binding, LtgA may require additional stimuli to initiate PG degradation. We hypothesized that, similar to certain antibiotics targeting PG synthases, which induce wild-type cells to degrade PG and form pseudospores in rich liquid media (Zhang et al., 2023b, O’Connor & Zusman, 1997), glycerol may alter cellular metabolism, leading to reduced PG synthesis. If this is the case, cells with upregulated PG synthesis should remain rods even in the presence of glycerol.

To test our hypothesis, we used the vanillate-inducible promoter (Iniesta et al., 2012) to overexpress murA as a merodiploid in the wild-type background. Because MurA catalyzes the first committed step of PG synthesis that produces UDP-MurNAc (Egan et al., 2020, Rohs & Bernhardt, 2021), elevated MurA levels are expected to channel more cellular resources toward PG synthesis. Overproduction of MurA in the presence of 200 µM vanillate resulted in a heterogeneous cell population, with normal cells coexisting alongside elongated ones (Figure 2A, B, Figure 2 – source data 1). Similar to the cells that overexpressed LtgA (Figure 4B), this heterogeneity likely reflects variable murA induction or the unsynchronized growth stages within the population. Overall, excessive MurA increased the average length of vegetative cells by 19.6%. Cells overproducing MurA also displayed heterogeneity during glycerol-induced sporulation. While some cells transitioned into spheres, many retained their rod shape even after 6 h of induction (Figure 2A, B, Figure 2 – source data 1). Notably, sporadic lysis of rod-shaped cells began 2 h post-induction and became increasingly frequent with continued incubation (Figure 2A). Potentially, elevated muropeptide production may lead to the accumulation of toxic intermediates that the relatively insufficient LTGs failed to degrade (Weaver et al., 2022). Nevertheless, the capacity of MurA overexpression to enable certain cells to circumvent glycerol-induced sporulation implies that PG production acts as a regulatory cue for PG degradation.

To test if excessive MurA can also reduce PG degradation during slow sporulation, we induced MurA overexpression on solid CF agar containing 100 µM vanillate. After 96 h of incubation, these cells developed fruiting body-like aggregates (Figure 3A). To test if such fruiting bodies contained mature spores, we scraped them from submerged CF medium and tested the germination rate of spores on CYE agar after sonication. The fruiting bodies formed by cells overexpressing MurA yielded 7.6% of the colonies produced by an equivalent number of wild-type cells, indicating that upregulated PG synthesis also prevents slow sporulation. In contrast, cells grown without vanillate progressed normally through glycerol-induced sporulation and starvation-induced fruiting body formation, showing no differences compared to wild-type cells (Figure 3B – Figure supplement 2). Therefore, the sporulation of M. xanthus is sensitive to PG synthesis and increased PG synthesis balances PG degradation in both sporulation pathways.

Discussion

Our findings demonstrate that M. xanthus, a non-firmicute bacterium, relies on PG degradation to change cell shape during sporulation. This mechanism presents a significant challenge for sporulating cells: preserving their structural integrity while simultaneously dismantling the primary framework that upholds it. Our findings revealed that besides regulating the production of LTGs, cells could control the pace of PG degradation in a rapid sporulation pathway through the regulation between two LTGs and hence allow spore maturation.

Studying the regulation between two LTGs is especially challenging both in vivo and in vitro for a few reasons. First, it is difficult to monitor their instantaneous activities in vivo. Secondly, muropeptide analysis may not distinguish their roles because their products bear the same signature, anhydro-MurNAc. Thirdly, some enzymes require special substrates or activators, which would be difficult to study in vitro. To overcome these technical hurdles, we leveraged single-particle mobility to quantify the PG- binding of these enzymes. The simultaneous occurrence of reduced LtgA mobility and PG degradation during glycerol-induced sporulation indicates that the molecular dynamics of LtgA accurately mirrors its enzymatic activity. Such correlation between decreased particle mobility and increased enzymatic activity applies to many other PG- related enzymes, including multiple PG polymerases in E. coli and the endopeptidase DacB in M. xanthus (Lee et al., 2016, Zhang et al., 2023b, Yang et al., 2021). Because the size difference between PG and its related enzymes is a conserved feature in all bacteria, our method using single-particle tracking to quantify PG-binding can be applied in many organisms.

We discovered that LtgB, a slower-acting LTG, limits the access of the faster LtgA to the PG during the early phase of rapid sporulation, thereby regulating the rate of its degradation. Then how does LtgA breach the blockage of LtgB and gain access to PG as sporulation progresses? Because the overexpression of LtgA is sufficient to damage PG in uninduced cells (Figure 4B), we propose that the relative abundance of these two LTGs plays a critical role in determining the fate during fast sporulation. While LtgB nearly saturates PG binding sites in vegetative cells where few LtgA are produced, LtgA—upregulated after glycerol induction (Muller et al., 2010) —gradually outcompetes LtgB and thereby causes LtgB to disassociate from PG.

As increased PG synthesis enables many cells to bypass both sporulation pathways, it is reasonable to propose that diminished PG synthesis initiates PG degradation by LTGs. While functional coordination between PG synthesis and hydrolysis has been speculated for over 50 years (Koch, 1985, Koch, 1990), its mechanisms only came to light recently. Some endopeptidases are proposed to serve as the space-makers for PG synthases (Dorr et al., 2013, Singh et al., 2012). In M. xanthus, we reported that moenomycin, an antibiotic that specifically binds to the GTase domains of class A penicillin-binding proteins (aPBPs) (Sung et al., 2009, Lovering et al., 2007), specifically activates a PG endopeptidase DacB through PBP1a2, an aPBP (Zhang et al., 2023b). Because moenomycin mimics a growing glycan strand in the GTase domains of aPBPs, it actually locks aPBPs in their glycan-charged conformations (Sung et al., 2009, Lovering et al., 2007). Thus, similar to the moenomycin-bound form, glycan-charged PBP1a2 in physiological conditions recruits DacB to the PG assembly sites, which in turn, generates openings in the existing PG network as the crosslink sites for the TPase activity of PBP1a2. Such “make-before-break” mechanism (Koch & Doyle, 1985) could prevent unneeded hydrolysis and thus maintain PG integrity.

Compared to the well-characterized space-making functions of endopeptidases, the roles of LTGs, beyond PG recycling, remain ill-defined. Even less is known on their potential coordination with PG synthases (Weaver et al., 2023). Because LTGs from bacterial and phage origins are inhibited by L,D-crosslinks, they may functionally connect to PBPs that form D,D-crosslinks (Alvarez et al., 2024). Additionally, some LTGs, such as MltD and MltG in E. coli and MltG in Vibrio cholerae, prefer to degrade nascent, uncrosslinked glycan strands (Kaul et al., 2024, Weaver et al., 2022, Bohrhunter et al., 2021). Our recent report indicates that AgmT, an M. xanthus LTG homologous to E. coli MltG, detoxifies uncrosslinked glycan strands under antibiotic stress and modifies crosslinked PG scaffolds to attach a motility machinery to PG (Ramirez Carbo et al., 2024). Then how does reduced PG synthesis activate LtgA in M. xanthus? Although MltE, the closest homolog of LtgA in E. coli, degrades glycan strands regardless of their crosslinking status or peptide content, it shows a marked preference for uncrosslinked substrates (Dik et al., 2017, Fibriansah et al., 2012). Similar to AgmT and MltE, LtgA might also bind to both crosslinked and uncrosslinked glycan strands. Consistent with its high mobility, LtgA may coordinate with diffusive PG polymerases and cleave the uncrosslinked glycan strands. During rapid sporulation, as nascent glycan strands decline due to muropeptide depletion, LtgA increasingly binds to the crosslinked PG scaffold, which is evidenced by its reduced mobility. This binding intensifies further as PG degradation continues, establishing a positive feedback loop. In this case, LtgB serves as a brake that dampens this positive feedback, moderates PG degradation, and thus allows sporulating cells to maintain integrity by assembling polysaccharide coats. Such cross-regulation between PG hydrolases, which is currently under investigation, provides yet another mechanism by which bacteria dynamically maintain their PG cell walls.

Materials and Methods

Bacterial strains and growth conditions

Vegetative M. xanthus cells were grown in liquid CYE medium (10 mM MOPS pH 7.6, 1% (w/v) Bacto™ casitone (BD Biosciences), 0.5% yeast extract and 8 mM MgSO4) at 32 °C, in 125-ml flasks with vigorous shaking, or on CYE plates that contains 1.5% agar. We used strain DZ2 as the wild-type M. xanthus strain (Campos & Zusman, 1975). Knock-out mutants were constructed by electroporating DZ2 cells with 4 µg of plasmid DNA. Transformed cells were plated on CYE plates supplemented with 100 mg/ml sodium kanamycin sulfate or 10 mg/ml tetracycline. The strains and plasmids used in this study are listed in Table 1.

Sporulation and spore purification

Cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2. Glycerol was added to 1 M to induce sporulation. The liquid culture was incubated at 32 °C with vigorous shaking. After 24 h, spores and cells were harvested by centrifugation (10 min, 10,000 g and 25 °C). To further purify the wild-type spores, the pellet was resuspended in 5 ml water, the remaining vegetative cells were eliminated by sonication (Cole Palmer 4710 ultrasonic homogenizer, 30% output, 10 cycles), and sonication-resistant spores were washed three times with water and collected by centrifugation (5 min, 10,000 g and 4 °C). The elimination of vegetative cells was confirmed by DIC microscopy.

For phenotypic assays on starvation-induced fruiting body formation, vegetative cells (10 ul), at a concentration of 4 × 109 ml-1, were spotted on CF (0.015% Casitone, 0.2% sodium citrate, 0.1% sodium pyruvate, 0.02% (NH4)2SO4, 10 mM MOPS (pH 7.6), 8 mM MgSO4 and 1 mM KH2PO4) plates containing an agar concentration of 1.5%, incubated at 32°C. To purify fruiting body spores, cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2 and harvested by centrifugation (10 min, 10,000 g and 25 °C). The pellet was resuspended in 1 ml water and plated on two 150 mm CF plates containing 1.5% agar using a spreader. The plates were air-dried and incubated at 32 °C for 120 h. For the wild-type and ΔltgA strains, fruiting bodies were scraped from the plates, suspended in 2 ml water, and subjected to sonication (Cole Palmer 4710 ultrasonic homogenizer, 60% output, 20 cycles). Spores were washed three times in water and collected by centrifugation (5 min, 10,000 g and 4 °C). For the ΔltgB strains, the sonication step was omitted.

Quantification of starvation-induced spores

We used the submerge starvation culture to quantify starvation-induced spores. Cells were first grown in liquid CYE medium overnight and adjusted to OD600 1 using fresh CYE. 50 ml of the asjusted culture was then poured to a Ø 150 mm polystyrene Petri dish and incubated without shaking for 24 h at 32 °C. Liquid and unattached cells were poured out carefully before adding 50 ml liquid CF medium. Petri dished were incubated without shaking for 96 h at 32 °C. Finally, liquid and unattached cells were poured out carefully and the cells attached to the bottom of Petri dished were scraped and resuspened into 1 ml water. The suspensions were then sonicated to eliminate vegetative cells, diluted and plated on solid CYE medium. Sonication-resistant spores were then numerated by counting the colonies after a 96-h incubation at 32 °C. Data were collected using three biological replicates.

Peptidoglycan purification and UPLC analysis

For PG analysis, samples were processed as previously described for Gram negative bacteria (Alvarez et al., 2016, Desmarais et al., 2013). Vegetative cells were harvested at mid-stationary phase by centrifugation (30 min, 8,000 g). Vegetative cells and purified spores (as described in the previous section) were resuspended and boiled in 1x PBS with 5% SDS for 2 h. The sacculi were repeatedly washed with MilliQ water by ultracentrifugation (150,000 g, 10 min, 20 °C) to remove the remaining SDS. The samples were then treated with muramidase (100 µg/mL) for 16 hours at 37 °C. Muramidase digestion was stopped by boiling and coagulated proteins were removed by centrifugation (10 min, 15,000 g). The supernatants were first adjusted to pH 8.5-9.0 with sodium borate buffer and then sodium borohydride was added to a final concentration of 10 mg/ml. After reducing the samples at room temperature for 30 min, the pH was adjusted to pH 3.5 with orthophosphoric acid.

UPLC analyses of muropeptides were performed on a Waters UPLC system (Waters Corp., USA) equipped with an ACQUITY UPLC BEH C18 Column (Waters Corp., USA) and a dual wavelength absorbance detector. Elution of muropeptides was detected at 204 nm. Muropeptides were separated at 45 °C using a linear gradient from buffer A (0.1% formic acid in water) to buffer B (0.1% formic acid in acetonitrile) in a 25- min run, with a 0.30 ml/min flow.

Muropeptide identification

Muropeptide identity was confirmed by MS and MS/MS analysis on a Xevo G2/XS Q- TOF mass spectrometer (Waters Corp., USA) coupled to the UPLC system with the same column, solvents, and gradients as mentioned above. Data acquisition and processing were performed using the UNIFI software package (Waters Corp., USA). Muropeptides were assigned based on: (i) accurate mass matching to theoretical monoisotopic masses of expected M. xanthus PG building blocks (Bui et al., 2009, White et al., 1968) and (ii) comparison of retention times with those reported in previous analysis with similar PG compositions. For vegetative cells analysis, relative total PG amounts were calculated by comparison of the total intensities of the chromatograms (total area) from three biological replicas normalized to the same initial biomass and extracted with the same volumes. Quantification of muropeptides was based on their relative abundances (relative area of the corresponding peak).

Imaging and data analysis

For all imaging experiments on isolated spores/cells, we spotted 5 μl of cells grown in liquid CYE medium to OD600 ∼1 on agar (1.5%) pads and imaged using a Andor iXon Ultra 897 EMCCD camera (effective pixel size 160 nm) on an inverted Nikon Eclipse- Ti™ microscope with a 100✕ 1.49 NA TIRF objective. Fruiting bodies were photographed after a 120-h incubation on CF agar using a Nikon SMZ1000 microscope and an OMAX A3590U digital camera. Cell morphology was imaged at indicated time points using DIC microscopy. The geometric aspect ratios (L/W) of spores/cells were determined using a custom algorithm written in MATLAB, which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b).

For sptPALM, M. xanthus cells were grown in CYE to 4 ×108 ml, spotted on 1.5% agar pads and subjected to highly inclined and laminated optical sheet (HILO) illumination (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013, Tokunaga et al., 2008). PAmCherry was activated using a 405-nm laser (0.3 kW/cm2), excited and imaged using a 561-nm laser (0.2 kW/cm2). Images were acquired at 10 Hz. For each sptPALM experiment, single PAmCherry particles were localized in at least 100 individual cells from three biological replicates. sptPALM data were analyzed using a MATLAB (MathWorks) script, which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b, Ramírez Carbó & Nan, 2026). Briefly, cells were identified using differential interference contrast images. Single PAmCherry particles inside cells were fit by a symmetric 2D Gaussian function, whose center was assumed to be the particle’s position (Fu et al., 2018). Particles that explored areas smaller than 160 nm × 160 nm (within one pixel) in 0.4 - 1.2 s were considered immotile (Fu et al., 2018, Zhang et al., 2023a, Zhang et al., 2023b). D of all the mobile particles was determined from a linear fit to the first four points of the MSD using a formula MSD = 4DΔt (Fu et al., 2018, Lee et al., 2016). Error bars were the standard derivation of 1,000 bootstrap samples using the published method (Morgenstein et al., 2015). Sample trajectories were generated using the TrackMate (Ershov et al., 2022) plugin in the ImageJ suite (https://imagej.net).

Immunofluorescence

Cell samples were resuspended and boiled in 1x PBS with 5% SDS for 2 h. The sacculi were repeatedly washed with water by centrifugation (5 min, 15,000 g and 25 °C) to remove the remaining SDS. Remaining proteins were further removed by incubation with proteinase K (1 µg/ml) for 2 h at 37 °C. Microscope cover slides were prepared by coating with 100 µl 0.1% poly-L-lysine for 5 min at room temperature (RT), washing with ddH2O and air drying. 100 µl samples were spotted onto the slides and incubated for 20 min at RT. The slides were then washed three times with phosphate buffer saline (PBS; 81 mM Na2HPO4, 15 mM KH2PO4, 1.37 M NaCl, 27 mM KCl, pH 7.4), and then incubated with anti-PG primary antibody16 diluted 1:1000 in PBS at RT for 1 h. Slides were washed 6 times with PBS and then incubated with Alexa Fluor 488 conjugated goat-anti-rabbit secondary antibody (Invitrogen) at a 1:1000 dilution in PBS at RT for 1 h in the dark. Slides were washed 6 times with PBS. Each cover glass was applied to a supporting slide, fixed with nail polish, and stored at 4 °C, if necessary. Images were recorded using a 488-nm laser.

Protein expression and purification

DNA sequences encoding amino acids 21 – 243 of LtgA and 26 - 710 of LtgB were amplified by polymerase chain reaction (PCR) and inserted into the pET28a vector (Novagen) between the restriction sites of EcoRI and HindIII and used to transform E. coli strain BL21(DE3). Transformed cells were cultured in 20 ml LB (Luria-Bertani) broth at 37 °C overnight and used to inoculate 1 L LB medium supplemented with 1.0% glucose. Protein expression was induced by 0.1 mM IPTG (isopropyl-h-d- thiogalactopyranoside) when the culture reached an OD600 of 0.8. Cultivation was continued at 16 °C for 10 h before the cells were harvested by centrifugation at 6,000 × g for 20 min and lysed by sonication in buffer A (20 mM Tris-HCl pH 8.0, 200 mM NaCl), (Nan et al., 2010, Nan et al., 2006). Proteins were loaded to an NGC™ Chromatography System (BIO-RAD) and 5-ml HisTrap™ columns (Cytiva) and eluted by buffer B (20 mM Tris-HCl pH 8.0, 200 mM NaCl, 500 mM immidazole) (Pogue et al., 2018, Nan et al., 2010). Purified proteins were concentrated using Amicon™ Ultra centrifugal filter units (Millipore Sigma) with a 10-kDa molecular weight cutoff and stored at −80 °C.

LTG activity (RBB) assay

PG was purified following the published method (Zhang et al., 2023b, Alvarez et al., 2016, Ramirez Carbo et al., 2024). In brief, M. xanthus cells were grown until mid- stationary phase and harvested by centrifugation (6,000 × g, 20 min, 25 °C). Supernatant was discarded and the pellet was resuspended and boiled in 1× PBS with 5% SDS for 2 h. SDS was removed by repetitive wash with water and centrifugation (21,000 × g, 10 min, 25 °C). Purified PG from 100 ml culture was suspended into 1 ml 1× PBS and stored at −20 °C. RBB labelling of PG was performed essentially as previously described (Uehara et al., 2010, Jorgenson et al., 2014). Purified sacculi were incubated with 20 mM RBB in 0.25 M NaOH overnight at 37 °C. Reactions were neutralized by adding equal volumes of 0.25 M HCl and RBB-labeled PG was collected by centrifugation at 21,000 × g for 15 min. Pellets were washed repeatedly with water until the supernatants became colorless. RBB-labelled sacculi were incubated with purified LtgA or LtgB (1 mg/ml) at 25 °C for 12 h. lysozyme (1 mg/ml) was used as a positive control. Dye release was quantified by the absorption at 595 nm from the supernatants after centrifugation (21,000 × g, 10 min, 25 °C).

Immunoblotting

Cells were grown in liquid CYE medium overnight, harvested, and concentrated to OD600 10. 10 µl of cell suspension was mixed with 10 µl of loading buffer and 10 µl of the mixture was subjected to SDS-PAGE. The expression and stability of PAmCherry- labeled proteins were determined by immunoblotting using an anti-mCherry antibody (Rockland Immunochemicals, Inc., Lot 46705) and a goat anti-Rabbit IgG (H+L) secondary antibody, HRP (Thermo Fisher Scientific, catalog # 31460). MreB was detected as the loading control using an anti-MreB serum(Mauriello et al., 2010) and the same secondary antibody. The blots were developed with Pierce™ ECL Western Blotting Substrate (Thermo Fisher Scientific REF 32109) and a MINI-MED 90 processor (AFP Manufacturing).

eLife Assessment

This important study elucidates the roles of two lytic transglycosylase (LTG) enzymes in the progression of spore formation in the research model species Myxococcus xanthus. Consistent with previous work indicating loss of peptidoglycan during the laboratory-induced rapid spore formation process, the new data demonstrate a similar loss of peptidoglycan during the normal, slow sporulation process. Solid evidence is provided for the roles of the two LTGs in spore formation and for interplay between the functions of these enzymes and peptidoglycan synthetic systems. These findings may have broader implications for understanding PG metabolism across a range of bacterial species.

Reviewer #2 (Public review):

The authors initial goal was to demonstrate loss of PG during the slow sporulation process of Myxococcus xanthus, with examination of the PG degradation products in order to implicate possible enzymes involved. Upon finding a predominance of LTG products, they examined sporulation in strains lacking each of the 14 candidate LTGs encoded in the genome, leading to the identification of two sporulation-linked LTGs. An extensive characterization of the roles played by these LTGs. One LTG is responsible for the slow sporulation PG degradation, while another is required for the rapid sporulation process. Interestingly, the "slow" LTG seems to provide an important regulatory brake on the rapid enzyme. Single molecule fluorescent tracking of these enzymes was used to develop a model for their interaction with PG that mimics their observed activity. The rate of PG synthesis activity was also shown to impact the rate of PG degradation, suggesting potential interplay between the synthetic and degradative enzymes.

Strengths:

The genetic analysis to identify sporulation-linked LTGs and their effects on growth sporulation, and spore properties was well done and productive. The fluorescence microscopy to track LTG mobility, presumably tied to activity, produced a convincing argument about the mechanism of regulation of one LTG by another. The authors have responded well to most points of the previous review.

Weaknesses:

While the impact of LTGs on sporulation was clearly demonstrated, the PG analysis that resulted in the study of LTGs raised some important unanswered questions. The analyses suggest that the PG is degraded to quite small fragments, which would normally be lost during the purification of PG. The conclusions concerning the PG degradation during sporulation needs to be clarified, as described below. The authors suggest a "new mechanism of sporulation" when they have actually simply identified an important factor (PG degradation by LTGs) within a complex "process of sporulation". This needs to be reflected also in title of the paper.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #2 (Public review):

The authors initial goal was to demonstrate loss of PG during the slow sporulation process of Myxococcus xanthus, with examination of the PG degradation products in order to implicate possible enzymes involved. Upon finding a predominance of LTG products, they examined sporulation in strains lacking each of the 14 candidate LTGs encoded in the genome, leading to the identification of two sporulation-linked LTGs. An extensive characterization of the roles played by these LTGs. One LTG is responsible for the slow sporulation PG degradation, while another is required for the rapid sporulation process. Interestingly, the "slow" LTG seems to provide an important regulatory brake on the rapid enzyme. Single molecule fluorescent tracking of these enzymes was used to develop a model for their interaction with PG that mimics their observed activity. The rate of PG synthesis activity was also shown to impact the rate of PG degradation, suggesting potential interplay between the synthetic and degradative enzymes.

Strengths:

The genetic analysis to identify sporulation-linked LTGs and their effects on growth sporulation, and spore properties was well done and productive. The fluorescence microscopy to track LTG mobility, presumably tied to activity, produced a convincing argument about the mechanism of regulation of one LTG by another. The authors have responded well to most points of the previous review.

Weaknesses:

While the impact of LTGs on sporulation was clearly demonstrated, the PG analysis that resulted in the study of LTGs raised some important unanswered questions. The analyses suggest that the PG is degraded to quite small fragments, which would normally be lost during the purification of PG. The conclusions concerning the PG degradation during sporulation needs to be clarified, as described below. The authors suggest a "new mechanism of sporulation" when they have actually simply identified an important factor (PG degradation by LTGs) within a complex "process of sporulation". This needs to be reflected also in title of the paper.

We have addressed the reviewer’s concerns and updated the text.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

(1) Lines 100-125: I am still concerned about clarity in the description of the muropeptides from spores. The claim is that 90% of the recovered muropeptides are anhydro LTG products. If LTGs had truly cleaved so many of the glycosidic bonds such that 90% of the muramic acid was now in the anhydro form, then all of the PG in the spores would be in small fragments (dimers and trimers containing 4-6 sugars and 8-12 amino acids), and these would be soluble and lost during purification. Whatever the form of the PG, it must not be easily soluble, so it must be either larger than that or bound to something else. The fact that the muropeptides are solubilized by muramidase indicates that there are some NAM-NAG bonds remaining, and cleavage of these should release some non-anhydro muropeptides. The spore muropeptide chromatograms have a large, late "mound" of UV-absorbing material that was released by muramidase digestion, and two of the identified anhydro products are present in this mound. It is not clear how these two anhydro products were quantified within this mound and what other (presumably) muropeptide species might be present in this mound. Some explanation of how these two muropeptides were quantified is needed.

We thank the reviewer for this careful point. As the reviewer notes, our purification procedure recovers only sedimentable PG, and any PG fragments solubilized by LTG activity would be lost during the washing steps and therefore not represented in our analysis. We have now explicitly acknowledged this limitation and discussed its implications in the revised text:

“Consequently, any PG fragments solubilized by LTG activity during sporulation would be lost at this stage, and the muropeptides we detect derive from the more highly crosslinked material that survives the procedure.

“Despite the overall decline in identified muropeptides, anhydro-muropeptides a minor component of vegetative PG were enriched in both spore types (Figure 1).”

I feel that the authors need to address some of this uncertainty in the results and discussion. They might be able to say that anhydro-muropeptides represent 90% of the "identified muropeptides" but need to acknowledge that there is a great deal of unidentified material released by the muramidase digestion. The data might also indicate that the LTG activity solubilizes much of the PG, which is lost, and results in recovery of only highly cross-linked muropeptides that might survive the PG purification process.

We agree that two of the identified anhydro species elute within a broad, unresolved region of the chromatogram that accounts for a substantial fraction of the muramidase-released material. Because reliable assignment and quantification of individual components within this region is not possible, we concluded that expressing anhydro-muropeptides as a fraction of the total identified muropeptides could be misleading. We have therefore removed the quantitative statement and now describe the enrichment of anhydromuropeptides in spores relative to vegetative cells qualitatively:

“The identified muropeptides, however, represent only a part of the material released by muramidase from spores: a substantial, as a late-eluting portion of the chromatogram could not be assigned, and its composition remains unknown.”

This would not eliminate the conclusion that "their abundance in spores indicates that certain LTGs must play essential roles (perhaps change to "might play important roles") in M. xanthus sporulation", which leads to the remaining studies in the paper.

Following the reviewer's suggestion, we have softened the conclusion of this section to state that LTGs "may play important roles" in sporulation.

(2) The authors have changed the statement about a "new mechanism of sporulation" at the beginning of the discussion, but this language is still in the paper title. Something more like "Programmed peptidoglycan degradation plays an important role in Myxococccus sporulation"

Following the reviewer's suggestion, we changed the title to “A novel mechanism for morphological change during bacterial sporulation based on programmed peptidoglycan degradation”.

Version 3

Abstract

Many bacteria form spores to endure unfavorable conditions. While Firmicutes generate endospores through cell division, sporulation in non-Firmicutes remains less understood. The Gram-negative bacterium Myxococcus xanthus undergoes sporulation through two distinct mechanisms: rapid sporulation triggered by chemical induction and slow sporulation driven by starvation, both occurring independently of cell division. Instead, these processes depend on the complete degradation of the peptidoglycan (PG) cell wall by two lytic transglycosylases (LTGs), LtgA and LtgB. Remarkably, LtgB programs the pace of PG degradation by LtgA during rapid sporulation, ensuring a controlled process that prevents abrupt PG breakdown and the formation of non-resistant pseudospores. In addition to regulation between LTGs, PG degradation is also influenced by its synthesis; cells exhibiting increased muropeptide production often circumvent sporulation. These findings not only reveal novel mechanisms of bacterial sporulation but also shed light on the regulatory network governing PG dynamics.

Introduction

Spores are metabolically dormant cells that can survive unfavorable conditions, including extremes of temperature, desiccation, and ionizing radiation (Huang & Hull, 2017, Hutchison et al., 2014). Sporulation, the process of spore development, is a strategy utilized by a wide variety of organisms, from bacteria and protozoa to fungi and plants. In addition to protective spore coats, spore resilience relies on cytosol dehydration and DNA compaction—key processes that decrease cell volume and alter cell shape (Higgins & Dworkin, 2012). Thus, morphological differentiation is a hallmark of sporulation. In bacterial spore formers, as peptidoglycan (PG) cell walls largely determine cell morphology, their sporulation requires profound PG remodeling.

PG is a rigid, mesh-like macromolecule that is composed of glycan strands of repeating units of N-acetyl glucosamine (GluNAc)-N-acetyl muramic acid (MurNAc) crosslinked by peptides (Egan et al., 2020). PG encloses the entire cytoplasmic membrane, and its rigidity provides mechanical support against osmotic stress. For this reason, PG is an essential structure for most bacteria and a major target for antibacterial treatments. Sporulation provides invaluable opportunities for understanding the dynamics of PG, which is controlled by multiple synthases and hydrolases. PG synthases include glycosyltransferases (GTases) that polymerize glycan strands and transpeptidases (TPases) that form peptide crosslinks (Egan et al., 2020). PG hydrolases, also known as autolysins, include glycosidases that break the glycan strands, and amidases and endo/carboxypeptidases that cleave the peptide crosslinks (Egan et al., 2020, Rohs & Bernhardt, 2021, van Heijenoort, 2011).

Firmicutes, including Gram-positive bacteria like Bacilli and Clostridia, produce endospores. Their morphological transition from rod-shaped vegetative cells to ovoid spores occurs through asymmetric cell division, resulting in the formation of a smaller forespore and a larger mother cell. Eventually, the forespore becomes an oval endospore after being engulfed by the mother cell (Higgins & Dworkin, 2012, McKenney & Eichenberger, 2012). A mature endospore contains two PG layers: the germ cell wall, derived from the forespore’s original PG, and a thickened PG cortex deposited by the mother cell (Popham & Bernhards, 2015). While a few Gram-negative spore formers also belong to the Firmicutes phylum, they share conserved sporulation genes with Bacilli and Clostridia, which suggests similar sporulation mechanisms (Tocheva et al., 2011, Yutin & Galperin, 2013).

Sporulation by non-firmicutes bacteria has been largely overlooked. Myxobacteria, a group of Gram-negative, non-firmicutes spore formers, were first assigned to the phylum δ-proteobacteria, but recently reclassified into the newly established phylum Myxococcota (Waite et al., 2020). Myxococcus xanthus is a model organism of myxobacteria. Rod-shaped M. xanthus cells can undergo sporulation via two distinct pathways, both resulting in spherical spores (Kroos et al., 2025). First, in response to certain chemical signals, such as glycerol and dimethyl sulfoxide (DMSO), individual M. xanthus cells can rapidly transition into isolated spherical spores in aqueous environments within hours (Dworkin & Gibson, 1964). Second, millions of cells can aggregate on solid starvation media and develop into spore-filled fruiting bodies, a process that takes a few days to complete (O’Connor & Zusman, 1988). Both sporulation pathways are programmed, tightly controlled processes that involve over 1,000 genes (Muller et al., 2010, Munoz-Dorado et al., 2019). In contrast to endospore formation, cell division is not involved in either the M. xanthus sporulation mechanisms, and M. xanthus lacks the homologs of the sporulation genes in Firmicutes (Aramayo & Nan, 2022). Thus, sporulating M. xanthus must accomplish the rod-to-sphere transition through yet to be discovered PG remodeling mechanisms, which provide invaluable opportunities to investigate PG dynamics (Zhang et al., 2021).

Distinct from Bacilli and Clostridia, M. xanthus spores lack cortex PG (Bui et al., 2009, Zhang et al., 2020, Voelz & Dworkin, 1962). Rather, their resistance is derived from polysaccharide coats (Kottel et al., 1975, Perez-Burgos et al., 2020, Saidi et al., 2021, Muller et al., 2012). Sporulating cells that fail to deposit coat polysaccharides on their surfaces produce spherical pseudospores that lack resistance (Wartel et al., 2013, Zhang et al., 2020, Holkenbrink et al., 2014). In a pioneering study, Bui et al. did not detect muropeptides in glycerol-induced spores, indicating that M. xanthus degrades its PG during rapid sporulation (Bui et al., 2009). However, it remains unclear whether starvation-induced spores within fruiting bodies still retain PG and how M. xanthus orchestrates programmed morphological changes.

In this report, we used ultra-performance liquid chromatography (UPLC) to demonstrate that M. xanthus degrades PG in both sporulation pathways. Through mutagenesis studies, we identified LtgA and LtgB, two lytic transglycosylases (LTGs) essential for these sporulation pathways. Remarkably, LtgB regulates the pace of PG degradation by LtgA during rapid sporulation, preventing the formation of non-resistant pseudospores due to abrupt PG breakdown. This research not only uncovers novel mechanisms of sporulation in non-firmicutes but also highlights the crucial role of cross-regulation between PG hydrolases in maintaining cell integrity.

Results

PG is degraded in both spore types

To investigate if starvation-induced spores in M. xanthus fruiting bodies retain PG, we broke the fruiting bodies after 120 h of starvation and purified saculli from spores. For comparison, we also purified the sacculi from vegetative cells and glycerol-induced spores. The purification procedure yields only sedimentable PG, as all soluble fragments are removed during the washing steps. Consequently, any PG fragments solubilized by LTG activity during sporulation would be lost at this stage, and the muropeptides we detect derive from the more highly crosslinked material that survives the procedure. The resulting sacculi were then digested with muramidase, and the solubilized muropeptides were analyzed by UPLC (see Materials and Methods). The two spore types showed similar profiles of a discernible presence of muropeptides that resembled those found in vegetative cells, albeit in significantly reduced quantities (Figure 1). Importantly, the most abundant muropeptides identified in the vegetative cells, MurNAc-tetrapeptide monomer (M4) and MurNAc-tetrapeptide dimer (D44), were nearly absent in both spore types (Figure 1). Such residual muropeptides in both spore types are inadequate for forming continuous PG layers (Bui et al., 2009). This observation indicates that spore development in both pathways involves the breakdown of the vegetative PG cell wall.

Mature M. xanthus spores induced by either glycerol or starvation do not contain significant muropeptides.

UPLC muropeptides profiles of vegetative cells, glycerol- and starvation-induced spores indicate that the major muropeptides species, M4 and D44, in vegetative cells are diminished in both spore types, while spores are enriched in anhydro-muropeptides (Anh). Only muropeptides that could be confidently identified by MS/MS analysis are labelled. The characteristic peaks are labeled as follows: M, monomeric muropeptide (uncrosslinked); D, dimeric muropeptide (crosslink connecting two muropeptides), T, trimeric muropeptide (crosslink connecting three muropeptides). Numbers refer to the status of the peptide side chain (3, tripeptide; 4, tetrapeptide). Red characters mark the muropeptides only detected in spores.

Despite the overall decline in identified muropeptides, anhydro-muropeptides — a minor component of vegetative PG — were enriched in both spore types (Figure 1). Especially, anhydro-MurNAc-tetrapeptide trimer (T444Anh) that was only detected in trace amounts in vegetative cells, became a prominent muropeptide in both spore types (Figure 1). Moreover, new anhydro-muropeptides, including anhydro, anhydro-MurNAc-tetrapeptide dimer (D44Anh,Anh), anhydro, anhydro-MurNAc-tetrapeptide trimer (T444Anh,Anh), and an anhydrodimer without a NAG (D44Anh(-NAG)), were only present in spores (Figure 1). The identified muropeptides, however, represent only a part of the material released by muramidase from spores: a substantial, as a late-eluting portion of the chromatogram could not be assigned, and its composition remains unknown. Because anhydro-muropeptides are the signature products of LTGs (Dik et al., 2017, Williams et al., 2018), their abundance in spores indicates that certain LTGs may play important roles in M. xanthus sporulation.

M. xanthus sporulation requires two LTGs

The genome of M. xanthus encodes 14 putative LTGs (Aramayo & Nan, 2022, Ramirez Carbo et al., 2024). We imaged the cells of the 14 knockout mutants (Ramirez Carbo et al., 2024) after 6 h of glycerol induction and identified two mutants that displayed abnormality in sporulation (Figure 2, Figure 2 – figure supplement 1). We then imaged the wild-type and mutant cells at different time points after glycerol induction and used their length/width (L/W) ratios to monitor potential defects in the sporulation process. Wild-type cells initiated sporulation within 1 h of glycerol induction, which was evidenced by the increase of cell width and decrease of cell length (Figure 2A). Consistent with the change in cell morphology, their L/W ratios decreased continuously and stabilized after 2 h of induction, when sporulation completed. In contrast, the cells that carry the deletion of K1515_20820 (MXAN_RS16290) (Aramayo & Nan, 2022) were able to shorten cell length slightly but retained rod shape after prolonged induction (Figure 2B, Figure 2 – source data 1).

LtgA and LtgB are required for glycerol-induced sporulation, whereas MurA overexpression inhibit sporulation.

A) Bright field images of cells at different time points after glycerol-induction. Black arrows point to lysing cells. B) Quantitative analysis of glycerol-induced sporulation using the length/width ratio (L/W) of cells. Whiskers indicate the 25th - 75th percentiles and red dots the median. The total number of cells analyzed is shown on top of each plot. C) LtgA and LtgB are homologous to E. coli MltE. The asterisk marks the conserved active site. Other putative LTGs do not affect either sporulation pathway significantly, which is shown in Figure 2 – figure supplement 1.

Surprisingly, another mutant that carries the deletion of K1515_17460 (MXAN_RS19615) started to lose rod shape immediately after glycerol induction and became spherical within 1 h (Figure 2A, 2B, Figure 2–source data 1). However, different from the wild-type spores that appeared dark and heterogeneous under differential interference contrast (DIC) microscopy, the spheres formed by this mutant appeared bright and homogenous, similar to the pseudospores from the ΔaglQS mutant that lack the motor for depositing spore coat polysaccharides onto cell surfaces (Zhang et al., 2020) (Figure 2A). To test whether these spheres are real spores, we subjected them to sonication and quantified their survival rate using a Helber bacterial counting chamber. After sonication, while 91 ± 6% (calculated from three independent experiments, n > 1,000) of the wild-type spores appeared intact, only 3 ± 1% of the spheres produced by the K1515_17460 deletion mutant remained. Thus, this mutant indeed formed pseudospores that lacked the resistance against sonication. As the products of K1515_20820 and K1515_17460 show homology to MltE, an LTG encoded by Escherichia coli emtA (49.0% and 38.4% sequence identity, respectively) (Figure 2C), we named them ltgA and ltgB, respectively. Both LtgA and LtgB are required for forming resistant spores via the rapid sporulation pathway, albeit playing different roles in the rod-to-sphere transition.

To determine if these LTGs are also required for starvation-induced sporulation, we grew them in rich liquid media and spotted cells on solid starvation (CF) media. After 96 h of incubation, both mutants produced dark fruiting bodies on the agar surface that were comparable to those of the wild-type (Figure 3A). To test if these mutants form starvation-induced spores, we incubated cells in static liquid CF medium using polystyrene Petri dishes for 96 h, scraped the cells attached to the dishes, suspended them in 1 ml water, subjected them to sonication, then plated them on solid rich media. After 96 h of incubation, while the wild-type fruiting bodies produced colonies of 3389 ± 1141 cfu/ml (n = 3), the ΔltgA and ΔltgB cells failed to form colonies after five days of incubation (Figure 3B, Figure 3 – source data 1), indicating that both LtgA and LtgB are essential for forming starvation-induced spores.

LtgA, LtgB, and MurA regulate starvation-induced sporulation.

A) The absence of LtgA and LtgB, and the overexpression of MurA do not affect the formation of fruiting body– like aggregates on starvation agar. B) Both LtgA and LtgB are essential for forming starvation-induced spores, while MurA overexpression negatively regulates this process. Sonication-resistant spores were counted as colony formation units (cfu) after 96 h of incubation on CYE agar. Data are shown as mean ± standard deviation (n = 3).

LtgA and LtgB play distinct roles in different sporulation pathways

Do LtgA and LtgB degrade PG during sporulation? To answer this question, we purified cell sacculi and used immunofluorescence and an anti-PG serum (de Pedro et al., 1997) to visualize the remaining PG. The sacculi of vegetative cells from the wild-type, ΔltgA, and ΔltgB strains were not visible under DIC microscopy, likely due to their flattened structures minimizing light interference. However, PG from all three strains was readily detected in the fluorescence channel (Figure 4A). After 6 h of glycerol-induction, the sacculi of both the wild-type and ΔltgA cells remained visible under DIC microscopy (Figure 4A), likely due to the deposition of spore coat polysaccharides that sustained unflattened cell structures (Wartel et al., 2013, Holkenbrink et al., 2014). After the sacculus purification process, only background PG signals were detected in glycerol-induced wild-type spores (Figure 4A). In contrast, the ΔltgA cells retained PG in substantial quantities (Figure 4A). Sacculi from the ΔltgB pseudospores still contained PG but showed lower fluorescence intensity (Figure 4A). While the remaining PG sacculi of these pseudospores were largely spherical, they lost integrity during purification, with many sacculi displaying irregular shapes in the fluorescence channel (Figure 4A). Thus, compared to LtgA, the enzymatic activity of LtgB plays a more limited role in PG degradation during chemical-induced sporulation. Collectively, LtgB appears to be a pace-keeper that prevents abrupt PG degradation. Different from the induced wild-type and ΔltgA cells, sacculi from the ΔltgB pseudospores were undetectable under DIC microscopy (Figure 4A), reflecting their flattened shapes. This phenotype suggests that further acceleration of the morphological transitions associated with rapid sporulation may impede the formation of a rigid polysaccharide coat.

LtgA and LtgB play distinct roles in the two sporulation pathways.

A) While LtgA is required for PG degradation during glycerol-induced sporulation, LtgB is the major LTG for forming starvation-induced spores. PG was detected using an anti-PG serum and a fluorescence-conjugated secondary antibody. PG sacculi were purified from cells after 6 h and 120 h of glycerol-induced and starvation-induced sporulation, respectively. Flattened sacculi are not visible under bright field microscopy. White arrows point to the sacculi in irregular shapes. BF, bright field. B) The overexpression of LtgA, but not LtgB, collapses rod-shape in vegetative cells. C) Purified LtgA and LtgB solubilize dye-labeled PG sacculi at different rates. Lysozyme and buffer serve as the positive and negative controls, respectively. Absorption at 595 nm was measured after 18 h incubation at 25 ⁰C. Data are presented as mean values ± SD from three technical replicates. The inset shows purified LtgA and LtgB in a Coomassie stained gel. Scale bars, 5 μm.

To determine the roles of LtgA and LtgB in slow sporulation, we first scraped the wild-type fruiting bodies from CF agar surface (Figure 2C), dispersed spores by sonication, purified their sacculi, and visualized PG using immunofluorescence. Spherical spores from the wild-type cells were visible under DIC microscope, indicating that they retained unflattened shapes. These fruiting body spores lacked PG-specific fluorescence (Figure 4A), consistent with their markedly reduced PG content (Figure 1). However, we cannot exclude the possibility that the polysaccharide spore coats (Voelz & Dworkin, 1962) hinder antibody access to PG. Second, we investigated if ΔltgA and ΔltgB cells degraded PG during starvation-induced sporulation. Because the aggregates formed by ΔltgA and ΔltgB cells on CF agar did not contain mature spores, we scrapped cell aggregates from agar surface and purified their sacculi without sonication. The ΔltgA cells formed spherical, spore-like cells void of PG (Figure 4A), indicating that while LtgB is sufficient for PG degradation during slow sporulation, LtgA is only required for spore maturation. In contrast, most of the ΔltgB cells were still rod-shaped, indistinguishable from vegetative ones (Figure 4A). These observations indicate that LtgB is an essential LTG for PG degradation during slow sporulation.

The two sporulation pathways vary greatly in duration: glycerol-induced rapid sporulation is completed within four hours, while starvation-induced slow sporulation unfolds over several days. Based on this distinction, we hypothesized that LtgA and LtgB could degrade PG at different rates. We used a vanillate-inducible promoter (Iniesta et al., 2012) to overexpress LtgA and LtgB as a merodiploids. Upon induction with 200 µM vanillate, cells overproducing LtgA exhibited heterogeneous morphology, likely resulting from variations in LtgA expression or differences in cell cycle stages within the population. Many cells lost rod shape even in the absence of glycerol, indicating that these cells over-degraded their PG by LtgA (Figure 4B). In contrast, the overexpression of LtgB using the same method did not affect the morphology of vegetative cells (Figure 4B). To test if LtgB overexpression was induced by vanillate, we added a photo-activatable mCherry (PAmCherry) tag to the overexpression constructs of LtgA and LtgB. As shown in Figure 4 – figure supplement 1, in the presence of vanillate, the expression of both LTGs increased significantly. These findings suggest that LtgB exhibits very low LTG activity in vegetative cells, as overexpression of the protein fails to induce substantial PG degradation.

To further investigate the activities of these LTGs, we expressed the periplasmic domains of wild-type LtgA (amino acids 21 – 243) and LtgB (amino acids 26 – 710) in E. coli (Figure 4C, Figure 4 – source data 1). We purified PG from wild-type M. xanthus cells, labeled it with Remazol brilliant blue (RBB), and tested if the purified LTGs hydrolyze labeled PG in vitro and release the dye (Jorgenson et al., 2014, Ramirez Carbo et al., 2024, Uehara et al., 2010). Wild-type LtgA solubilized dye-labeled PG, which absorbed light at 595 nm, demonstrating stronger hydrolytic activity than lysozyme, an enzyme that specifically cleaves β-1,4-glycosidic bonds in PG. In contrast, PG incubated with LtgB only showed minimum release of the dye, indicating slow PG hydrolysis (Figure 4C, Figure 4 – source data 1). While we cannot rule out the possibility that our purification and reaction conditions were suboptimal for LtgB, both the in vitro RBB assay and the in vivo phenotype resulting from LtgB overexpression support our hypothesis that during vegetative growth, LtgB is less efficient in PG degradation than LtgA.

LtgB regulates LtgA during glycerol-induced sporulation

Consistent with its role in PG degradation, in a microarray-based transcriptome analysis, ltgA transcription was found to increase about twofold during rapid sporulation (4 h) but remain unchanged during slow sporulation (96 h) in the closely related DK1622 strain (Muller et al., 2010). Conversely, ltgB expression gradually rises during slow sporulation, reaching 1.8 times the vegetative level at 96 h, while remaining stable during rapid sporulation (Muller et al., 2010, Munoz-Dorado et al., 2019). However, transcriptomic data alone do not account for the opposing roles of LtgA and LtgB in the rod-to-sphere transition during rapid sporulation. The resemblance in morphology between glycerol-induced ΔltgB cells and uninduced cells overproducing LtgA (Figure 4B) suggests that the slower-acting LtgB may offset the rapid activity of LtgA during sporulation, allowing sufficient time for spore maturation.

To test if LtgB regulates LtgA, we constructed a ΔltgA ΔltgB double deletion strain. Cells from this strain shortened their length slightly but failed to abolish rod shape after prolonged glycerol induction, phenocopying the sporulation defect of the ΔltgA strain (Figure 2A, 2B, Figure 2 – source data 1). Hence, LtgB is a regulator upstream of LtgA.

LtgB exhibits a response to glycerol induction earlier than LtgA

Biochemical reactions on PG are unique for the stark size difference between the enzymes and their substrates. While the enzymes are in nanometer scales, their substrates, the PG sacculi, span several micrometers. Under the microscope, PG remains stationary but PG-related enzymes are free to move. Even for M. xanthus that moves on surfaces, PG-related enzymes move at least two orders of magnitude faster than the cell/PG (Zhang et al., 2023b, Ramirez Carbo et al., 2024). Thus, when diffusive enzymes bind to PG, their mobility decreases (Lee et al., 2016, Zhang et al., 2023b). For instance, DacB, another PG hydrolase, reduces its single-particle mobility in the conditions where its activity is activated (Zhang et al., 2023b). By tracking single fluorescently-labeled enzyme particles, we can approximate their PG-binding in different physiological conditions and genetic backgrounds (Ramirez Carbo et al., 2024, Zhang et al., 2023b, Ramírez Carbó & Nan, 2026). We individually expressed PAmCherry-labeled LtgA and LtgB using their native loci and promoters. Both labeled LTGs accumulated as full-length proteins (Figure 5 – figure supplement 1A) and the PAmCherry tags did not negatively affect either glycerol or starvation-induced sporulation pathways, indicating that these fusion proteins were fully functional (Figure 5 – figure supplement 1B, C). Notably, vegetative cells produce significantly less LtgA than LtgB, consistent with the RNAseq data in which ltgA transcription was below the detection limit (Munoz-Dorado et al., 2019) (Figure 5 – figure supplement 1A).

LtgB regulates the PG-binding of LtgA during glycerol-induced sporulation.

A) LtgB exhibits a response to glycerol induction earlier than LtgA. Representative trajectories of LtgA and LtgB before (uninduced) and after (1 min and 30 min) glycerol induction. The overall distribution of both LTGs is displayed using the composite of 100 consecutive frames taken at 100-ms intervals. Single-particle trajectories of PAmCherry were generated from the same frames. Individual trajectories are distinguished by colors. Scale bars, 5 μm. B and C) the absence of LtgB reduces the diffusion of LtgA, which is reflected in the increase of immobile population (B) and the decrease in diffusion coefficient (D) (C), and these effects are especially prominent in the cells before (uninduced) and immediately after (1 min) glycerol induction. For each protein and condition, particles were identified from at least 100 cells and three independent experiments. The total number of particles analyzed is shown on top of each plot. Error bars were the standard derivation of 1,000 bootstrap samples and * indicates a difference of > 0.005.

We used a 405-nm excitation laser (0.3 kW/cm2, 0.1 s) to activate the fluorescence of a few labeled LTG particles randomly in each cell and quantified their localization using a 561-nm laser at 10 Hz using single particle tracking photo-activated localization microscopy (sptPALM, see Materials and Methods) (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013, Ramírez Carbó & Nan, 2026). As free PAmCherry particles diffuse extremely fast in the cytoplasm, entering and exiting the focal plane frequently, they usually appear as blurry objects that cannot be followed at 10 Hz close to the cell surface (Fu et al., 2018, Zhang et al., 2023b). For this reason, the noise from free PAmCherry due to potential protein degradation was negligible.

Single-particles of PAmCherry-labeled LtgA and LtgB can be categorized into two populations, immobile and mobile. The immobile particles remained within a single pixel (160 nm × 160 nm) before photobleach, and the mobile ones displayed typical diffusion (Figure 5A, Figure 5 – source data 1). For the particles that switched between mobile and immobile states, our algorithm categorized them as mobile and calculated their diffusion coefficients (D) from their entire trajectories that contained both mobile and immobile segments. Hence, binding to PG not only increases the immobile population of the enzyme particles but also decreases the D of the mobile particles. In vegetative cells where large-scale PG degradation does not occur, 23.9% (n = 1175) and 18.2% (n = 824) of LtgA and LtgB particles were immobile, respectively (Figure 5B, Figure 5 – source data 1). D values of the mobile population were (2.62 ± 0.20) × 10-2 µm2/s (n = 894) for LtgA-PAmCherry and (2.80 ± 0.25) × 10-2 µm2/s (n = 674) for LtgB-PAmCherry (Figure 5C, Figure 5 – source data 1).

We then determined how the dynamics of both LTGs respond to glycerol induction in the rapid sporulation pathway. Immediately (1 min) after adding glycerol, the dynamics of LtgA remained little changed, when 21.0% (n = 928) of particles were immobile and the D of the mobile ones was (2.53 ± 0.19) × 10-2 µm2/s (n = 733) (Figure 5B, C, Figure 5 – source data 1). In contrast, the mobility of LtgB decreased significantly, with the immobile population increased to 28.0% (n = 1359) and D of the mobile population decreased to (2.17 ± 0.28) × 10-2 µm2/s (n = 978) (Figure 5B, C, Figure 5 – source data 1). Therefore, these results suggest that upon glycerol induction, LtgB rapidly enhances its binding to PG.

We then chose 30 min after glycerol induction as a time point for rapid PG degradation, which is reflected by the dramatic decrease of L/W during the first hour of sporulation (Figure 2A, B, Figure 2 – source data 1). Compared to its inert response 1 min after glycerol induction, the mobility of LtgA particles decreased significantly at 30 min, when the immobile population increased to 31.5% (n = 1163) and D of the mobile population decreased to (1.80 ± 0.21) × 10-2 µm2/s (n = 797) (Figure 5B, C, Figure 5 – source data 1). The reduced mobility of LtgA aligns with its function as the primary LTG in glycerol-induced sporulation. Strikingly different from LtgA, as sporulation advanced to the 30-min mark, LtgB’s mobility returned to its pre-induction level, with the immobile population decreased to 10.2% (n = 1549) and D of the mobile population increased to (2.87 ± 0.28) × 10-2 µm2/s (n = 1391) (Figure 5B, C, Figure 5 – source data 1). Therefore, the PG-binding of LtgB shows a negative correlation with PG degradation. Taken together, LtgA and LtgB display opposite responses to glycerol induction, in which LtgB rapidly binds to PG before yielding to LtgA, whose PG-binding is concurrent with the rod-to-sphere transition.

LtgB blocks LtgA from binding PG in the early stage of glycerol-induced sporulation

Does LtgB’s early PG-binding suppress PG degradation by LtgA? To answer this question, we expressed LtgA-PAmCherry using the native ltgA locus and promoter in the ΔltgB background. In vegetative cells, the absence of LtgB significantly reduced the mobility of single LtgA-PAmCherry particles, suggesting that LtgB does affect LtgA’s binding to the PG (Figure 5B, C, Figure 5 – source data 1). Similarly, we expressed LtgB-PAmCherry using the native ltgB locus and promoter in the ΔltgA background. In contrast, the mobility of single LtgB-PAmCherry particles only decreased slightly in the absence of LtgA (Figure 5B, C, Figure 5 – source data 1). These results suggest that while LtgB significantly reduces LtgA’s binding to PG, probably due to its higher expression level, LtgA has only a modest impact on LtgB’s ability to bind PG.

Strikingly different from its slow response to glycerol in wild-type cells, LtgA increased its PG-binding immediately (1 min) after glycerol induction in the ΔltgB background, confirming that LtgB may compete with LtgA for access to PG in the early stage of sporulation (Figure 5B, C, Figure 5 – source data 1). In contrast, at 30 min of induction, a time point when LtgB dissociated from PG, its absence no longer affected LtgA’s PG-binding (Figure 5B, C, Figure 5 – source data 1). Similar to the observation in wild-type cells, the lack of LtgA did not affect the PG-binding of LtgB at either 1 min or 30 min after glycerol induction (Figure 5B, C, Figure 5 – source data 1). In summary, the sequential PG-binding by these two LTGs controls the pace of PG-degradation during glycerol-induced sporulation. LtgB, a slow LTG, regulates the progression of rapid sporulation by preventing LtgA, a fast LTG, from excessively degrading PG, thereby ensuring spore maturation.

Upregulated PG synthesis negatively affects PG degradation

Since LtgA is less mobile in vegetative ΔltgB cells, i.e. strongly binds to PG (Figure 5B, C, Figure 5 – source data 1), why do these cells retain their rod shape rather than losing it spontaneously before glycerol induction? In addition to PG-binding, LtgA may require additional stimuli to initiate PG degradation. We hypothesized that, similar to certain antibiotics targeting PG synthases, which induce wild-type cells to degrade PG and form pseudospores in rich liquid media (Zhang et al., 2023b, O’Connor & Zusman, 1997), glycerol may alter cellular metabolism, leading to reduced PG synthesis. If this is the case, cells with upregulated PG synthesis should remain rods even in the presence of glycerol.

To test our hypothesis, we used the vanillate-inducible promoter (Iniesta et al., 2012) to overexpress murA as a merodiploid in the wild-type background. Because MurA catalyzes the first committed step of PG synthesis that produces UDP-MurNAc (Egan et al., 2020, Rohs & Bernhardt, 2021), elevated MurA levels are expected to channel more cellular resources toward PG synthesis. Overproduction of MurA in the presence of 200 µM vanillate resulted in a heterogeneous cell population, with normal cells coexisting alongside elongated ones (Figure 2A, B, Figure 2 – source data 1). Similar to the cells that overexpressed LtgA (Figure 4B), this heterogeneity likely reflects variable murA induction or the unsynchronized growth stages within the population. Overall, excessive MurA increased the average length of vegetative cells by 19.6%. Cells overproducing MurA also displayed heterogeneity during glycerol-induced sporulation. While some cells transitioned into spheres, many retained their rod shape even after 6 h of induction (Figure 2A, B, Figure 2 – source data 1). Notably, sporadic lysis of rod-shaped cells began 2 h post-induction and became increasingly frequent with continued incubation (Figure 2A). Potentially, elevated muropeptide production may lead to the accumulation of toxic intermediates that the relatively insufficient LTGs failed to degrade (Weaver et al., 2022). Nevertheless, the capacity of MurA overexpression to enable certain cells to circumvent glycerol-induced sporulation implies that PG production acts as a regulatory cue for PG degradation.

To test if excessive MurA can also reduce PG degradation during slow sporulation, we induced MurA overexpression on solid CF agar containing 100 µM vanillate. After 96 h of incubation, these cells developed fruiting body-like aggregates (Figure 3A). To test if such fruiting bodies contained mature spores, we scraped them from submerged CF medium and tested the germination rate of spores on CYE agar after sonication. The fruiting bodies formed by cells overexpressing MurA yielded 7.6% of the colonies produced by an equivalent number of wild-type cells, indicating that upregulated PG synthesis also prevents slow sporulation. In contrast, cells grown without vanillate progressed normally through glycerol-induced sporulation and starvation-induced fruiting body formation, showing no differences compared to wild-type cells (Figure 3B – Figure supplement 2). Therefore, the sporulation of M. xanthus is sensitive to PG synthesis and increased PG synthesis balances PG degradation in both sporulation pathways.

Discussion

Our findings demonstrate that M. xanthus, a non-firmicute bacterium, relies on PG degradation to change cell shape during sporulation. This mechanism presents a significant challenge for sporulating cells: preserving their structural integrity while simultaneously dismantling the primary framework that upholds it. Our findings revealed that besides regulating the production of LTGs, cells could control the pace of PG degradation in a rapid sporulation pathway through the regulation between two LTGs and hence allow spore maturation.

Studying the regulation between two LTGs is especially challenging both in vivo and in vitro for a few reasons. First, it is difficult to monitor their instantaneous activities in vivo. Secondly, muropeptide analysis may not distinguish their roles because their products bear the same signature, anhydro-MurNAc. Thirdly, some enzymes require special substrates or activators, which would be difficult to study in vitro. To overcome these technical hurdles, we leveraged single-particle mobility to quantify the PG-binding of these enzymes. The simultaneous occurrence of reduced LtgA mobility and PG degradation during glycerol-induced sporulation indicates that the molecular dynamics of LtgA accurately mirrors its enzymatic activity. Such correlation between decreased particle mobility and increased enzymatic activity applies to many other PG-related enzymes, including multiple PG polymerases in E. coli and the endopeptidase DacB in M. xanthus (Lee et al., 2016, Zhang et al., 2023b, Yang et al., 2021). Because the size difference between PG and its related enzymes is a conserved feature in all bacteria, our method using single-particle tracking to quantify PG-binding can be applied in many organisms.

We discovered that LtgB, a slower-acting LTG, limits the access of the faster LtgA to the PG during the early phase of rapid sporulation, thereby regulating the rate of its degradation. Then how does LtgA breach the blockage of LtgB and gain access to PG as sporulation progresses? Because the overexpression of LtgA is sufficient to damage PG in uninduced cells (Figure 4B), we propose that the relative abundance of these two LTGs plays a critical role in determining the fate during fast sporulation. While LtgB nearly saturates PG binding sites in vegetative cells where few LtgA are produced, LtgA—upregulated after glycerol induction (Muller et al., 2010) —gradually outcompetes LtgB and thereby causes LtgB to disassociate from PG.

Although glycerol has been known for decades to trigger sporulation in M. xanthus, the underlying mechanism remains elusive. Notably, glycerol was recently found to disrupt peptidoglycan (PG) synthesis in B. subtilis; in contrast, this effect inhibits B. subtilis sporulation, which depends on enhanced PG synthesis (Updegrove et al., 2026). While elevated glycerol concentrations are associated with reduced PG synthesis in both species, whether and how glycerol affects PG synthesis in M. xanthus remains unclear. Furthermore, it is unknown whether this apparent correlation reflects a shared evolutionary origin or a common mechanistic basis. Nevertheless, PG dynamics must play critical roles in bacterial sporulation

As increased PG synthesis enables many M. xanthus cells to bypass both sporulation pathways, it is reasonable to propose that diminished PG synthesis initiates PG degradation by LTGs. While functional coordination between PG synthesis and hydrolysis has been speculated for over 50 years (Koch, 1985, Koch, 1990), its mechanisms only came to light recently. Some endopeptidases are proposed to serve as the space-makers for PG synthases (Dorr et al., 2013, Singh et al., 2012). In M. xanthus, we reported that moenomycin, an antibiotic that specifically binds to the GTase domains of class A penicillin-binding proteins (aPBPs) (Sung et al., 2009, Lovering et al., 2007), specifically activates a PG endopeptidase DacB through PBP1a2, an aPBP (Zhang et al., 2023b). Because moenomycin mimics a growing glycan strand in the GTase domains of aPBPs, it actually locks aPBPs in their glycan-charged conformations (Sung et al., 2009, Lovering et al., 2007). Thus, similar to the moenomycin-bound form, glycan-charged PBP1a2 in physiological conditions recruits DacB to the PG assembly sites, which in turn, generates openings in the existing PG network as the crosslink sites for the TPase activity of PBP1a2. Such “make-before-break” mechanism (Koch & Doyle, 1985) could prevent unneeded hydrolysis and thus maintain PG integrity.

Compared to the well-characterized space-making functions of endopeptidases, the roles of LTGs, beyond PG recycling, remain ill-defined. Even less is known on their potential coordination with PG synthases (Weaver et al., 2023). Because LTGs from bacterial and phage origins are inhibited by L,D-crosslinks, they may functionally connect to PBPs that form D,D-crosslinks (Alvarez et al., 2024). Additionally, some LTGs, such as MltD and MltG in E. coli and MltG in Vibrio cholerae, prefer to degrade nascent, uncrosslinked glycan strands (Kaul et al., 2024, Weaver et al., 2022, Bohrhunter et al., 2021). Our recent report indicates that AgmT, an M. xanthus LTG homologous to E. coli MltG, detoxifies uncrosslinked glycan strands under antibiotic stress and modifies crosslinked PG scaffolds to attach a motility machinery to PG (Ramirez Carbo et al., 2024). Then how does reduced PG synthesis activate LtgA in M. xanthus? Although MltE, the closest homolog of LtgA in E. coli, degrades glycan strands regardless of their crosslinking status or peptide content, it shows a marked preference for uncrosslinked substrates (Dik et al., 2017, Fibriansah et al., 2012). Similar to AgmT and MltE, LtgA might also bind to both crosslinked and uncrosslinked glycan strands. Consistent with its high mobility, LtgA may coordinate with diffusive PG polymerases and cleave the uncrosslinked glycan strands. During rapid sporulation, as nascent glycan strands decline due to muropeptide depletion, LtgA increasingly binds to the crosslinked PG scaffold, which is evidenced by its reduced mobility. This binding intensifies further as PG degradation continues, establishing a positive feedback loop. In this case, LtgB serves as a brake that dampens this positive feedback, moderates PG degradation, and thus allows sporulating cells to maintain integrity by assembling polysaccharide coats. Such cross-regulation between PG hydrolases, which is currently under investigation, provides yet another mechanism by which bacteria dynamically maintain their PG cell walls.

Materials and Methods

Bacterial strains and growth conditions

Vegetative M. xanthus cells were grown in liquid CYE medium (10 mM MOPS pH 7.6, 1% (w/v) Bacto™ casitone (BD Biosciences), 0.5% yeast extract and 8 mM MgSO4) at 32 °C, in 125-ml flasks with vigorous shaking, or on CYE plates that contains 1.5% agar. We used strain DZ2 as the wild-type M. xanthus strain (Campos & Zusman, 1975). Knock-out mutants were constructed by electroporating DZ2 cells with 4 µg of plasmid DNA. Transformed cells were plated on CYE plates supplemented with 100 mg/ml sodium kanamycin sulfate or 10 mg/ml tetracycline. The strains and plasmids used in this study are listed in Table 1.

Sporulation and spore purification

Cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2. Glycerol was added to 1 M to induce sporulation. The liquid culture was incubated at 32 °C with vigorous shaking. After 24 h, spores and cells were harvested by centrifugation (10 min, 10,000 g and 25 °C). To further purify the wild-type spores, the pellet was resuspended in 5 ml water, the remaining vegetative cells were eliminated by sonication (Cole Palmer 4710 ultrasonic homogenizer, 30% output, 10 cycles), and sonication-resistant spores were washed three times with water and collected by centrifugation (5 min, 10,000 g and 4 °C). The elimination of vegetative cells was confirmed by DIC microscopy.

For phenotypic assays on starvation-induced fruiting body formation, vegetative cells (10 ul), at a concentration of 4 × 109 ml-1, were spotted on CF (0.015% Casitone, 0.2% sodium citrate, 0.1% sodium pyruvate, 0.02% (NH4)2SO4, 10 mM MOPS (pH 7.6), 8 mM MgSO4 and 1 mM KH2PO4) plates containing an agar concentration of 1.5%, incubated at 32°C. To purify fruiting body spores, cells were grown in 25 ml liquid cell culture to OD600 0.8 – 1.2 and harvested by centrifugation (10 min, 10,000 g and 25 °C). The pellet was resuspended in 1 ml water and plated on two 150 mm CF plates containing 1.5% agar using a spreader. The plates were air-dried and incubated at 32 °C for 120 h. For the wild-type and ΔltgA strains, fruiting bodies were scraped from the plates, suspended in 2 ml water, and subjected to sonication (Cole Palmer 4710 ultrasonic homogenizer, 60% output, 20 cycles). Spores were washed three times in water and collected by centrifugation (5 min, 10,000 g and 4 °C). For the ΔltgB strains, the sonication step was omitted.

Quantification of starvation-induced spores

We used the submerge starvation culture to quantify starvation-induced spores. Cells were first grown in liquid CYE medium overnight and adjusted to OD600 1 using fresh CYE. 50 ml of the asjusted culture was then poured to a Ø 150 mm polystyrene Petri dish and incubated without shaking for 24 h at 32 °C. Liquid and unattached cells were poured out carefully before adding 50 ml liquid CF medium. Petri dished were incubated without shaking for 96 h at 32 °C. Finally, liquid and unattached cells were poured out carefully and the cells attached to the bottom of Petri dished were scraped and resuspened into 1 ml water. The suspensions were then sonicated to eliminate vegetative cells, diluted and plated on solid CYE medium. Sonication-resistant spores were then numerated by counting the colonies after a 96-h incubation at 32 °C. Data were collected using three biological replicates.

Peptidoglycan purification and UPLC analysis

For PG analysis, samples were processed as previously described for Gram negative bacteria (Alvarez et al., 2016, Desmarais et al., 2013). Vegetative cells were harvested at mid-stationary phase by centrifugation (30 min, 8,000 g). Vegetative cells and purified spores (as described in the previous section) were resuspended and boiled in 1x PBS with 5% SDS for 2 h. The sacculi were repeatedly washed with MilliQ water by ultracentrifugation (150,000 g, 10 min, 20 °C) to remove the remaining SDS. The samples were then treated with muramidase (100 µg/mL) for 16 hours at 37 °C. Muramidase digestion was stopped by boiling and coagulated proteins were removed by centrifugation (10 min, 15,000 g). The supernatants were first adjusted to pH 8.5-9.0 with sodium borate buffer and then sodium borohydride was added to a final concentration of 10 mg/ml. After reducing the samples at room temperature for 30 min, the pH was adjusted to pH 3.5 with orthophosphoric acid.

UPLC analyses of muropeptides were performed on a Waters UPLC system (Waters Corp., USA) equipped with an ACQUITY UPLC BEH C18 Column (Waters Corp., USA) and a dual wavelength absorbance detector. Elution of muropeptides was detected at 204 nm. Muropeptides were separated at 45 °C using a linear gradient from buffer A (0.1% formic acid in water) to buffer B (0.1% formic acid in acetonitrile) in a 25-min run, with a 0.30 ml/min flow.

Muropeptide identification

Muropeptide identity was confirmed by MS and MS/MS analysis on a Xevo G2/XS Q-TOF mass spectrometer (Waters Corp., USA) coupled to the UPLC system with the same column, solvents, and gradients as mentioned above. Data acquisition and processing were performed using the UNIFI software package (Waters Corp., USA). Muropeptides were assigned based on: (i) accurate mass matching to theoretical monoisotopic masses of expected M. xanthus PG building blocks (Bui et al., 2009, White et al., 1968) and (ii) comparison of retention times with those reported in previous analysis with similar PG compositions. For vegetative cells analysis, relative total PG amounts were calculated by comparison of the total intensities of the chromatograms (total area) from three biological replicas normalized to the same initial biomass and extracted with the same volumes. Quantification of muropeptides was based on their relative abundances (relative area of the corresponding peak).

Imaging and data analysis

For all imaging experiments on isolated spores/cells, we spotted 5 μl of cells grown in liquid CYE medium to OD600 ∼1 on agar (1.5%) pads and imaged using a Andor iXon Ultra 897 EMCCD camera (effective pixel size 160 nm) on an inverted Nikon Eclipse-Ti™ microscope with a 100✕ 1.49 NA TIRF objective. Fruiting bodies were photographed after a 120-h incubation on CF agar using a Nikon SMZ1000 microscope and an OMAX A3590U digital camera. Cell morphology was imaged at indicated time points using DIC microscopy. The geometric aspect ratios (L/W) of spores/cells were determined using a custom algorithm written in MATLAB, which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b).

For sptPALM, M. xanthus cells were grown in CYE to 4 ×108 ml, spotted on 1.5% agar pads and subjected to highly inclined and laminated optical sheet (HILO) illumination (Fu et al., 2018, Nan et al., 2015, Nan et al., 2013, Tokunaga et al., 2008). PAmCherry was activated using a 405-nm laser (0.3 kW/cm2), excited and imaged using a 561-nm laser (0.2 kW/cm2). Images were acquired at 10 Hz. For each sptPALM experiment, single PAmCherry particles were localized in at least 100 individual cells from three biological replicates. sptPALM data were analyzed using a MATLAB (MathWorks) script , which is available in the GitHub repository, https://github.com/NanLabMyxo/Rod_shape_paper (Zhang et al., 2020, Zhang et al., 2023a, Zhang et al., 2023b, Ramírez Carbó & Nan, 2026). Briefly, cells were identified using differential interference contrast images. Single PAmCherry particles inside cells were fit by a symmetric 2D Gaussian function, whose center was assumed to be the particle’s position (Fu et al., 2018). Particles that explored areas smaller than 160 nm × 160 nm (within one pixel) in 0.4 - 1.2 s were considered immotile (Fu et al., 2018, Zhang et al., 2023a, Zhang et al., 2023b). D of all the mobile particles was determined from a linear fit to the first four points of the MSD using a formula MSD = 4DΔt (Fu et al., 2018, Lee et al., 2016). Error bars were the standard derivation of 1,000 bootstrap samples using the published method (Morgenstein et al., 2015). Sample trajectories were generated using the TrackMate (Ershov et al., 2022) plugin in the ImageJ suite (https://imagej.net).

Immunofluorescence

Cell samples were resuspended and boiled in 1x PBS with 5% SDS for 2 h. The sacculi were repeatedly washed with water by centrifugation (5 min, 15,000 g and 25 °C) to remove the remaining SDS. Remaining proteins were further removed by incubation with proteinase K (1 µg/ml) for 2 h at 37 °C. Microscope cover slides were prepared by coating with 100 µl 0.1% poly-L-lysine for 5 min at room temperature (RT), washing with ddH2O and air drying. 100 µl samples were spotted onto the slides and incubated for 20 min at RT. The slides were then washed three times with phosphate buffer saline (PBS; 81 mM Na2HPO4, 15 mM KH2PO4, 1.37 M NaCl, 27 mM KCl, pH 7.4), and then incubated with anti-PG primary antibody16 diluted 1:1000 in PBS at RT for 1 h. Slides were washed 6 times with PBS and then incubated with Alexa Fluor 488 conjugated goat-anti-rabbit secondary antibody (Invitrogen) at a 1:1000 dilution in PBS at RT for 1 h in the dark. Slides were washed 6 times with PBS. Each cover glass was applied to a supporting slide, fixed with nail polish, and stored at 4 °C, if necessary. Images were recorded using a 488-nm laser.

Protein expression and purification

DNA sequences encoding amino acids 21 – 243 of LtgA and 26 - 710 of LtgB were amplified by polymerase chain reaction (PCR) and inserted into the pET28a vector (Novagen) between the restriction sites of EcoRI and HindIII and used to transform E. coli strain BL21(DE3). Transformed cells were cultured in 20 ml LB (Luria-Bertani) broth at 37 °C overnight and used to inoculate 1 L LB medium supplemented with 1.0% glucose. Protein expression was induced by 0.1 mM IPTG (isopropyl-h-d-thiogalactopyranoside) when the culture reached an OD600 of 0.8. Cultivation was continued at 16 °C for 10 h before the cells were harvested by centrifugation at 6,000 × g for 20 min and lysed by sonication in buffer A (20 mM Tris-HCl pH 8.0, 200 mM NaCl), (Nan et al., 2010, Nan et al., 2006). Proteins were loaded to an NGC™ Chromatography System (BIO-RAD) and 5-ml HisTrap™ columns (Cytiva) and eluted by buffer B (20 mM Tris-HCl pH 8.0, 200 mM NaCl, 500 mM immidazole) (Pogue et al., 2018, Nan et al., 2010). Purified proteins were concentrated using Amicon™ Ultra centrifugal filter units (Millipore Sigma) with a 10-kDa molecular weight cutoff and stored at -80 °C.

LTG activity (RBB) assay

PG was purified following the published method (Zhang et al., 2023b, Alvarez et al., 2016, Ramirez Carbo et al., 2024). In brief, M. xanthus cells were grown until mid-stationary phase and harvested by centrifugation (6,000 × g, 20 min, 25 °C). Supernatant was discarded and the pellet was resuspended and boiled in 1× PBS with 5% SDS for 2 h. SDS was removed by repetitive wash with water and centrifugation (21,000 × g, 10 min, 25 °C). Purified PG from 100 ml culture was suspended into 1 ml 1× PBS and stored at -20 °C. RBB labelling of PG was performed essentially as previously described (Uehara et al., 2010, Jorgenson et al., 2014). Purified sacculi were incubated with 20 mM RBB in 0.25 M NaOH overnight at 37 °C. Reactions were neutralized by adding equal volumes of 0.25 M HCl and RBB-labeled PG was collected by centrifugation at 21,000 × g for 15 min. Pellets were washed repeatedly with water until the supernatants became colorless. RBB-labelled sacculi were incubated with purified LtgA or LtgB (1 mg/ml) at 25 °C for 12 h. lysozyme (1 mg/ml) was used as a positive control. Dye release was quantified by the absorption at 595 nm from the supernatants after centrifugation (21,000 × g, 10 min, 25 °C).

Immunoblotting

Cells were grown in liquid CYE medium overnight, harvested, and concentrated to OD600 10. 10 µl of cell suspension was mixed with 10 µl of loading buffer and 10 µl of the mixture was subjected to SDS-PAGE. The expression and stability of PAmCherry-labeled proteins were determined by immunoblotting using an anti-mCherry antibody (Rockland Immunochemicals, Inc., Lot 46705) and a goat anti-Rabbit IgG (H+L) secondary antibody, HRP (Thermo Fisher Scientific, catalog # 31460). MreB was detected as the loading control using an anti-MreB serum(Mauriello et al., 2010) and the same secondary antibody. The blots were developed with Pierce™ ECL Western Blotting Substrate (Thermo Fisher Scientific REF 32109) and a MINI-MED 90 processor (AFP Manufacturing).